% Tree of Knowledge v3 --- rewritten following LoRA structural patterns
% Core claim: specialization depth, adapter rank, and storage tier are one dimension.

\documentclass{article}
\usepackage[preprint,nonatbib]{neurips_2024}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{hyperref}
\usepackage{url}
\usepackage{booktabs}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{nicefrac}
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage{tikz}
\usetikzlibrary{arrows.meta}
\usepackage{wrapfig}

\newcommand{\placeholder}[1]{\begin{center}\fbox{\parbox{0.85\textwidth}{\centering\textcolor{red}{\textbf{[FIGURE: #1]}}}}\end{center}}
\newcommand{\tok}{Tree of Knowledge}

\title{Tree of Knowledge: Differentiating Fluid Intelligence \\ from Modular Crystallized Knowledge in Language Models}

\author{
  Erik de Bruijn \\
  Independent Researcher \\
  \texttt{erik@erikdebruijn.nl}
}

\begin{document}
\maketitle

% ============================================================================
\begin{abstract}
We propose the \tok{} framework, which separates a language model's
\emph{fluid intelligence} (reasoning, syntax, attention patterns) from its
\emph{crystallized intelligence} (domain-specific knowledge) by placing them on
different parameter budgets. A shared dense core preserves reasoning capacity
while a hierarchy of variable-rank LoRA adapters---branching like a tree through
the model's depth---supplies domain knowledge on demand. Adapters are born at
rank-1 through \emph{learntropy-driven forking}: when a teacher detects bimodal
prediction difficulty in an expert, the expert splits into two children whose
divergence is accelerated by a contrastive loss. Rank grows only as the data
demands, yielding specialists measured in kilobytes rather than megabytes. We
ground the design in systematic negative results: six experimental arms on
Qwen3-1.7B show that copy-initialized experts trained on disjoint data remain
virtually identical (cosine similarity 0.998). LoRA initialization, disjoint
data, and contrastive pressure together reduce Phase~2 cosine similarity from
0.975 to 0.840 while matching the Phase~1 perplexity of 19.58 (baseline
21.20). The resulting architecture admits specialists that are trainable on
commodity hardware, composable by a community, and loadable from NVMe in
microseconds---with no inference-latency overhead.
\end{abstract}

% ============================================================================
\section{Introduction}
\label{sec:intro}

Current language models embed all knowledge---from common syntax to rare medical terminology---in the same parameters that perform reasoning. Adding knowledge means adding parameters, and therefore inference cost. The standard alternatives do not escape this trade-off: retrieval-augmented generation adds context tokens, tool calls add generation tokens, and longer contexts scale quadratically in attention. Every path to more knowledge costs inference budget.

The \tok{} framework largely eliminates this coupling by placing domain knowledge in separately-loadable adapters whose cost is zero until activated. The dense core handles reasoning and is always resident in VRAM. Crystallized knowledge lives in lightweight, hot-loadable LoRA adapters. Adding specialists does not slow the core's inference.

Prior work shows that learned models reside on a low intrinsic
dimension~\cite{li2018intrinsic,aghajanyan2021intrinsic}. We observe that
specialization depth, adapter rank, and storage tier are not three design
decisions---they are one. We distinguish two regimes:

\begin{itemize}[nosep,leftmargin=1em]
\item \textbf{Lens-like} (low rank, e.g.\ rank-1 to rank-4): the adapter does not alter the computation itself but \emph{redirects information} through the existing weights. The original FFN does the heavy lifting; the LoRA adapter bends the light. At 4--16\,KB, these corrections are always cached and cost nothing to load.
\item \textbf{Crystal-like} (high rank, e.g.\ rank-32 to rank-64): the $A \times B$ matrix is large enough to encode its own features---knowledge the original FFN does not possess. It is no longer a correction but an independent information carrier. At 128--256\,KB, these specialists are loaded from flash on demand.
\end{itemize}

The rank is not a hyperparameter to tune---it is a \emph{measurement} of how different this expert's knowledge is from the generalist core. Lens-like adapters say ``the same computation, slightly adjusted.'' Crystal-like adapters say ``fundamentally different knowledge lives here.''

\input{figure1.tex}

Standard Mixture-of-Experts (MoE) architectures attempt this separation but
fail in practice: load-balancing losses enforce near-uniform routing, preventing
any hot/cold distinction, and copy-initialized experts remain virtually
identical. We dedicate Section~\ref{sec:why-moe-fails} to this failure mode,
which motivates every design choice in \tok{}.

\paragraph{Concrete benefits.}
\begin{itemize}
    \item Specialists measured in \textbf{kilobytes}, not megabytes (4\,KB--256\,KB vs.\ 16\,MB per MoE expert).
    \item Trainable on \textbf{commodity hardware}, in parallel---each adapter trains independently.
    \item \textbf{Composable by a community}---contributors add specialists without retraining the core.
    \item \textbf{No inference-latency overhead}---the core runs at full speed; adapters merge into the forward pass.
\end{itemize}

Throughout this paper, we use \textbf{Qwen3-1.7B} (28 layers, 2.0B parameters)
as our running example, training on C4 data. All reported numbers are from this
model unless otherwise noted.

% ============================================================================
\section{Why Can't Standard MoE Differentiate?}
\label{sec:why-moe-fails}

The promise of Mixture-of-Experts is that different experts will learn different
things. In practice, they do---but not in the way we expect. Experts readily
specialize along \emph{functional} axes (structure vs.\ content, punctuation
vs.\ rare words) but resist specializing along \emph{domain} axes (medical
vs.\ legal vs.\ code). The word ``expert'' evokes a human domain specialist,
but MoE experts behave more like specialized tools: one handles
fasteners, another handles cutting, regardless of whether the project is
carpentry or plumbing. We spent six experimental arms and 40+ hours of GPU
training on Qwen3-1.7B discovering why domain specialization is harder than
functional specialization, and what it takes to cross that boundary.

\paragraph{The copy-initialization trap.}
The standard upcycling recipe~\cite{komatsuzaki2023sparse} creates experts by
copying the dense model's FFN weights. Each expert starts in the same basin of
attraction. Our decisive experiment: we forced completely disjoint data
partitions across experts, training for 12M tokens. Result: expert weight cosine
similarity remained at \textbf{0.998}. The copy initialization creates a basin
that disjoint data alone cannot escape on practical timescales.

\paragraph{Load-balancing destroys structure.}
MoE training adds an auxiliary load-balancing loss that penalizes uneven routing.
This is necessary to prevent expert collapse, but it has a side effect: it
enforces near-uniform token distribution across experts, eliminating the very
frequency structure that would allow hot/cold tiering. Every expert sees the
same distribution and converges to the same function.

\paragraph{Six arms, one conclusion.}
Table~\ref{tab:lessons} summarizes six experimental arms that systematically
explored alternatives. Prescriptive cost losses create routing structure (Gini
0.60) but the effect dilutes over training. Descriptive losses fail entirely
(Gini 0.05). Biological mechanisms help marginally (Gini 0.46). Without
explicit divergence pressure on the \emph{weights themselves}, no genuine
specialization emerges.

\begin{table}[t]
\centering
\caption{Six experimental arms on Qwen3-1.7B that establish what does \emph{not}
produce genuine expert differentiation.}
\label{tab:lessons}
\begin{tabular}{@{}llp{6.5cm}@{}}
\toprule
Arm & Key metric & Lesson \\
\midrule
A: Prescriptive cost loss  & Gini 0.60$\to$0.49 & Creates routing skew, but skew dilutes over training \\
B: Reuse-distance proxy     & Gini 0.05           & Descriptive loss produces no structure \\
C: Adaptive + cost          & 25 rounds, Gini dilutes & Piaget mechanism works, but routing Gini is wrong metric \\
D: Pure data-driven         & Gini 0.11           & No specialization without resource constraint \\
E: Bio-integrated           & Gini 0.46           & Six bio mechanisms help modestly, not fundamentally \\
F: KD-Warm upcycle          & CosSim 0.998        & Copy-init experts don't differentiate via data alone \\
\bottomrule
\end{tabular}
\end{table}

These failures taught us that expert differentiation requires three forces acting
\emph{together}: (1)~different starting points (not copy-initialization),
(2)~different training data, and (3)~explicit divergence pressure on the weights.
The \tok{} framework instantiates all three through learntropy-driven progressive
splitting with contrastive LoRA adapters.

% ============================================================================
\section{Method: Progressive LoRA Forking}
\label{sec:method}

The \tok{} architecture has three components: a shared trunk, a branching
mechanism, and a divergence loss. We describe each briefly; mathematical details
are in Appendix~\ref{app:math}.

\subsection{LoRA on FFN Layers, Not Attention}

A key design choice: we apply LoRA adapters to the \emph{feed-forward network}
(specifically \texttt{gate\_proj} and \texttt{up\_proj}) rather than to
attention layers. While Hu et al.~\cite{hu2022lora} demonstrated LoRA
primarily on attention projections, mechanistic interpretability research
suggests that FFN layers function as key-value memories that store factual
associations~\cite{geva2021transformer}, whereas attention layers perform
routing and compositional binding. Since our goal is to specialize
\emph{domain knowledge} (medical facts vs.\ legal facts vs.\ code patterns),
the FFN is the natural target: we are modifying what the model \emph{knows},
not how it \emph{routes} information. The attention layers remain shared across
all experts, preserving the model's general reasoning and compositional
abilities (fluid intelligence) while specialists modify only the crystallized
knowledge stored in the FFN.

% TODO: Update layer numbers when trunk-18 experiment validates the new boundary.
% Layer divergence analysis (2026-03-26) showed domain divergence begins at
% layer 17-18 (CKA 0.013 at layer 14 vs 0.106 at layer 18). The layer-14
% boundary below is from early experiments; trunk-18 experiment is running.
\subsection{Shared Trunk, Branching Depth}

For Qwen3-1.7B (28 layers), we partition the transformer into:
\begin{itemize}
    \item \textbf{Layers 1--14 (trunk):} Shared FFN, always VRAM-resident.
          This is the fluid intelligence core---general reasoning, syntax,
          attention patterns. $\sim$1.0B parameters.
    \item \textbf{Layers 15--21 (first fork):} 2--4 broad domain branches,
          each a LoRA adapter on the shared FFN ($\sim$2\,MB per branch).
    \item \textbf{Layers 22--28 (deep forks):} Up to 8+ branches
          ($\sim$0.5\,MB per branch).
\end{itemize}
Total specialist parameters: $\sim$6\,MB for 8 specialists, versus 800+\,MB for
8 full expert copies in standard MoE.

\subsection{Learntropy-Driven Mitosis}

% TODO: Learntropy definition is provisional. Wozniak (SuperMemo) independently
% coined "learntropy" as the attractiveness of a signal to the learn drive —
% a NET reward signal (reward minus decoding-failure penalty) that depends on
% prior knowledge and can go negative. Our current definition (raw CE loss) is
% a simplification: it does not distinguish "productively hard" (Wozniak's
% positive learntropy) from "impossibly hard" (negative learntropy). Bimodality
% detection partially recovers this distinction, but a proper learntropy
% function should model net learning value, not raw difficulty. We are working
% toward a definition closer to Wozniak's, operationalized as a loss function.
% See: https://supermemo.guru/wiki/Learntropy
%
\textbf{Definition: Learntropy.} We use \emph{learntropy} to denote the per-token prediction difficulty experienced by a specific expert:
\begin{equation}
\mathcal{L}_{\text{learn}}(e_i, t) = -\log P(t \mid \text{context}, e_i)
\end{equation}
where $e_i$ is expert $i$ and $t$ is the predicted token. The term was independently coined by Wozniak~\cite{wozniak2018learntropy} to denote the attractiveness of a signal to the learn drive system---a net reward that depends on prior knowledge and can go negative when material exceeds the learner's capacity. Our operationalization as raw cross-entropy loss is a simplification: it measures difficulty but does not distinguish productive difficulty (material in the learner's zone of proximal development~\cite{vygotsky1978}) from noise. Bimodality detection (below) partially recovers this distinction by identifying when an expert's loss distribution contains both learnable and unlearnable populations.

We use learntropy as a \emph{diagnostic signal for the learning system itself}: not only as a training objective but as an instrument that determines when structural change is needed. This reflects Shannon's insight that information content drives the allocation of representational resources~\cite{shannon1948}---and Piaget's observation that prediction failure (accommodation) triggers the creation of new cognitive structure~\cite{piaget1952}.

When an expert's learntropy distribution is bimodal---some tokens easy, others hard---the expert must either adapt or divide. Following biological cells, we distinguish two responses:

\textbf{Gene expression (rank increase).} The expert first attempts to handle both populations by increasing its rank---activating more parameters, like a cell expressing additional genes to handle a broader range of signals. This is the cheaper response: no routing overhead, no data partitioning.

\textbf{Mitosis (expert split).} If rank increase does not resolve the bimodality, the expert undergoes \emph{mitosis}: it divides into two daughter experts, each inheriting the parent's weights plus a small random perturbation. Like biological daughter cells, they start genetically identical but \emph{differentiate} through exposure to different environments (data partitions) and explicit pressure to diverge (contrastive loss). The procedure:
\begin{enumerate}
    \item Score each expert's tokens by learntropy.
    \item Test for bimodality (Hartigan's dip test).
    \item If bimodal and rank increase insufficient: duplicate the LoRA adapter.
          Each daughter inherits parent weights plus perturbation, starting at rank-1.
    \item Daughter experts \emph{differentiate} through their data environment and contrastive loss (Section~\ref{sec:contrastive}).
\end{enumerate}
The decision to divide is triggered by an information-theoretic signal, not a schedule---analogous to how cell division is triggered by growth signals, not a timer.

\subsection{Variable Rank as Specialization Measure}

After mitosis, each daughter expert begins at rank-1. Rank grows when reconstruction
error exceeds a threshold---the adapter needs more capacity to express what the
data demands. The rank \emph{is} the measurement of specialization depth, just as the number of active genes reflects a cell's degree of specialization:
\begin{itemize}
    \item \textbf{Rank 1--4} (4--16\,KB): A lens. Small correction, close to parent.
    \item \textbf{Rank 16--64} (64--256\,KB): A crystal. Fundamentally different
          computation.
\end{itemize}

\begin{table}[t]
\centering
\caption{Specialization depth, adapter rank, and storage tier are one dimension.}
\label{tab:unification}
\begin{tabular}{@{}llll@{}}
\toprule
Specialization & Adapter & Size & Storage tier \\
\midrule
Minimal (lens) & Rank-1 LoRA & 4\,KB & Always resident \\
Moderate       & Rank-16 LoRA & 64\,KB & VRAM cache \\
Substantial (crystal) & Rank-64 LoRA & 256\,KB & NVMe (instant load) \\
Full expert    & Dense copy & 12\,MB & NVMe (prefetchable) \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Contrastive Divergence Loss}
\label{sec:contrastive}

Without explicit pressure, siblings drift toward the shared optimum (we observed
CosSim 0.998). The contrastive loss provides the missing force:
\begin{equation}
\mathcal{L}_{\text{contrast}} = \lambda_c \sum_{i < j} \max\bigl(0,\; \cos(\theta_{ij}) - \tau\bigr)
\label{eq:contrastive}
\end{equation}
where $\theta_{ij}$ is the angle between sibling weight vectors and $\tau$ is a
target maximum similarity. Combined with disjoint data and LoRA initialization,
this produces the differentiation that eluded all six prior arms.

\subsection{Teacher-Guided Data Selection via Zone of Proximal Development}
\label{sec:zpd}

Contrastive loss differentiates expert \emph{weights}, but does not guarantee
that experts encode different \emph{knowledge}. Two experts can be orthogonal
in weight space yet both required for every input---as we observe with the
first-level Structure/Content split (Section~\ref{sec:hotloading}).

To produce modular experts whose removal causes \emph{selective} damage, each
expert must train on fundamentally different data. We select this data using a
teacher model and Vygotsky's Zone of Proximal Development (ZPD): the set of
examples that are learnable but not yet learned.

For each text chunk $x$ and expert $e_i$, we compute:
\begin{equation}
\text{ZPD}(x, e_i) = \frac{\text{PPL}_{\text{student}}(x, e_i)}{\text{PPL}_{\text{teacher}}(x)}
\label{eq:zpd}
\end{equation}
A high ZPD score means the student struggles where the teacher succeeds---the
knowledge exists in the teacher but is absent from the student. A low score
means either both struggle (noise, not knowledge) or both succeed (nothing to
learn). Each expert receives the chunks where \emph{its own} ZPD is highest,
producing genuinely disjoint training distributions.

Our hypothesis is that ZPD-guided data selection can produce domain-level
modularity that contrastive loss alone does not achieve
(Section~\ref{sec:hotloading}). This remains untested: we have scored a pool
of 50K chunks but have not yet trained experts on ZPD-selected data.

% ============================================================================
\section{Experiments}
\label{sec:experiments}

All experiments use Qwen3-1.7B on C4 data. We build incrementally: each
subsection adds one capability and reports the result.

\subsection{Layer Divergence Analysis: Where to Fork}
\label{sec:divergence}

Before placing adapters, we measure where domain-specific representations
emerge in the unmodified model. We pass 200 samples from each of five C4
domains (medical, code, legal, conversational, news) through Qwen3-1.7B and
compute per-layer Centered Kernel Alignment (CKA) between domain-specific
hidden state distributions.

\begin{table}[h]
\centering
\caption{Domain divergence by layer (selected). CKA measures alignment between
domain-specific hidden state distributions (higher = more domain-specific).}
\label{tab:divergence}
\begin{tabular}{@{}lrrr@{}}
\toprule
Layer range & CKA & Fisher ratio & Interpretation \\
\midrule
0--8 (early) & 0.003--0.004 & 0.004--0.005 & Domain-agnostic \\
9--14        & 0.005--0.013 & 0.006--0.009 & Minimal divergence \\
15--17       & 0.020--0.062 & 0.010--0.012 & Transition zone \\
\textbf{18--21} & \textbf{0.106--0.200} & \textbf{0.014--0.016} & \textbf{Domain-specific} \\
22--27       & 0.206--0.216 & 0.015--0.018 & Strong divergence \\
\bottomrule
\end{tabular}
\end{table}

CKA increases by an order of magnitude between layers 14 and 18 (0.013 to
0.106). This empirically determines the trunk boundary: layers below 18 carry
insufficient domain signal for expert specialization. Our initial experiments
(Sections~\ref{tab:phase1}--\ref{sec:hotloading}) used a trunk of 14 layers;
the layer divergence analysis motivated extending the trunk to 18 layers in
subsequent experiments.

\subsection{Baseline}

\begin{table}[h]
\centering
\caption{Qwen3-1.7B baseline on C4 validation.}
\label{tab:baseline}
\begin{tabular}{@{}lr@{}}
\toprule
Configuration & Perplexity \\
\midrule
Dense baseline (no adapters) & 21.20 \\
\bottomrule
\end{tabular}
\end{table}

The dense Qwen3-1.7B achieves a perplexity of 21.20 on the C4 validation set.
This is the number every subsequent experiment must match or improve.

\subsection{Phase 1: First Split with LoRA Adapters}

We freeze the trunk (layers 1--14) and attach rank-4 LoRA adapters to layers
15--28, training on full C4 data without domain partitioning. This validates
that the LoRA+trunk architecture can recover baseline quality.

\begin{table}[h]
\centering
\caption{Phase 1: LoRA adapters on the upper layers.}
\label{tab:phase1}
\begin{tabular}{@{}lrr@{}}
\toprule
Configuration & PPL & $\Delta$ from baseline \\
\midrule
Dense baseline & 21.20 & --- \\
Trunk + LoRA (layers 15--28) & 19.59 & $-$1.61 \\
\bottomrule
\end{tabular}
\end{table}

The LoRA adapters not only recover baseline quality but \emph{improve} it,
reducing perplexity to 19.59. The trunk preserves reasoning capacity while the
adapters specialize the upper layers.

\subsection{Phase 2: Contrastive Forking}

We create two sibling LoRA adapters at the first fork point, train them on
disjoint data partitions with the contrastive loss (Eq.~\ref{eq:contrastive}),
and measure differentiation.

\begin{table}[h]
\centering
\caption{Phase 2: Sibling differentiation with contrastive loss.}
\label{tab:phase2}
\begin{tabular}{@{}lrrr@{}}
\toprule
Configuration & CosSim & PPL & Note \\
\midrule
Disjoint data only (no contrast) & 0.975 & 19.62 & Minimal differentiation \\
Disjoint data + contrastive loss & 0.840 & 19.58 & Strong differentiation \\
\bottomrule
\end{tabular}
\end{table}

The contrastive loss reduces cosine similarity from 0.975 to 0.840---a
qualitative shift from ``nearly identical'' to ``measurably different
functions''---while \emph{improving} perplexity to 19.58. Differentiation does
not trade off against quality.

\subsection{Progressive Forking}
\label{sec:forking}

With the trunk boundary informed by the layer divergence analysis
(Section~\ref{sec:divergence}), we compare forking at two boundary points.

\begin{table}[h]
\centering
\caption{Trunk boundary comparison: Phase 2 results (2 experts, rank 4,
contrastive weight 0.1). Layer-18 trunk places experts in the domain-specific
zone identified by the divergence analysis.}
\label{tab:trunk-comparison}
\begin{tabular}{@{}lrrrr@{}}
\toprule
Configuration & Expert layers & CosSim & PPL & Params \\
\midrule
Trunk-14 (layers 15--28) & 14 & 0.840 & 19.58 & 1.9M \\
Trunk-14 + Phase 3 (rank 32) & 14 & 0.278 & 19.33 & 11.2M \\
Trunk-14 + Phase 3 (rank 4) & 14 & 0.433 & 19.45 & 1.9M \\
\textbf{Trunk-18 (layers 18--27)} & \textbf{10} & \textbf{0.771} & \textbf{20.14} & \textbf{655K} \\
Trunk-18 + Phase 3 (rank 4) & 10 & 0.276$^*$ & 19.87$^*$ & 655K \\
\bottomrule
\end{tabular}
\end{table}

\noindent $^*$\textit{Phase 3 interrupted at step 14K/48K; values are
intermediate, not final.}

The trunk-18 configuration achieves deeper differentiation (CosSim 0.276 at
rank 4) than trunk-14 at the same rank (0.433), with 3$\times$ fewer
parameters. Both fork location and rank constraint affect differentiation:
trunk-14 at rank 32 reaches 0.278, while trunk-14 at rank 4 reaches only
0.433. Trunk-18 at rank 4 matches the rank-32 result (0.276 vs.\ 0.278),
suggesting that \emph{placing experts in the domain-divergent zone compensates
for 8$\times$ fewer parameters}.

However, routing analysis (Section~\ref{sec:magnitude}) shows that none of
these configurations achieve domain-specific routing. Weight-space
differentiation (low CosSim) does not imply functional specialization. Recent
work on MoE expert roles~\cite{standing2026} confirms this pattern at scale:
a ``standing committee'' of experts handles functional/structural processing
across all domains, while domain-specific routing is peripheral.

No second fork (2$\to$4 experts) has triggered in any configuration on
generic C4 data.

\subsection{Hot-Loading Validation}
\label{sec:hotloading}

Our first hot-loading test \emph{failed}: removing either first-level expert
degraded all text types equally ($\sim$8\%). Token routing analysis
revealed Expert~0 handles rare/content words (71\% of rare tokens) while
Expert~1 handles punctuation and function words (86\% of punctuation). Every
sentence requires both structure and content---neither expert is independently
removable.

\textbf{Observation:} The first split is functional (structure vs.\ content),
not domain-specific. Both experts are always needed.

\textbf{Hypothesis (untested):} Domain-level modularity may emerge at deeper
tree levels, where sub-experts within the Content branch specialize on
different domains. If so, removing a domain sub-expert would selectively
degrade that domain's PPL. This requires a second fork, which has not yet
occurred in our experiments.

\textbf{Alternative explanations we cannot yet rule out:}
\begin{itemize}
    \item Deeper splits may also produce functional rather than domain
          differentiation, in which case the tree never becomes modular.
    \item The structure/content split may be an artifact of contrastive loss
          pushing toward the easiest orthogonal direction in weight space,
          not a meaningful linguistic distinction.
    \item The tree metaphor may not accurately describe the learning dynamics
          at all---the branching may reflect optimization landscape geometry
          rather than knowledge organization.
\end{itemize}

\placeholder{Table: Level-2 hot-loading experiment (pending second fork).
Remove a domain sub-expert, measure PPL per domain. Pre-registered prediction:
selective degradation on removed domain ($>$5\% PPL increase) with minimal
impact on others ($<$1\%). Failure threshold: uniform degradation similar to
level-1 ($\sim$8\% across all domains).}

\subsection{LoRA Magnitude Analysis}
\label{sec:magnitude}

To understand \emph{why} hot-loading fails, we measured the magnitude of LoRA
deltas relative to the trunk FFN output across 1000 evaluation tokens.

\begin{table}[h]
\centering
\caption{LoRA magnitude analysis at layers 14--27 (Phase 3 checkpoint,
rank-32). Pre-registered prediction: ratio $<$0.05 (small correction).
Result: ratio $\approx$0.25 (substantial correction).}
\label{tab:magnitude}
\begin{tabular}{@{}lrrr@{}}
\toprule
Metric & Expected & Measured & Verdict \\
\midrule
$\|\text{LoRA}\| / \|\text{FFN}\|$ (mean) & $<$0.05 & 0.249 & FAIL \\
Gate CosSim between experts & --- & 0.019 & Orthogonal \\
Routing selectivity (pref.\ vs non-pref.) & $>$2$\times$ & $<$5\% diff & FAIL \\
\bottomrule
\end{tabular}
\end{table}

The LoRA adapters contribute $\sim$25\% of the FFN output magnitude---they
are not small corrections but substantial transformations. Expert weight
matrices are nearly orthogonal (CosSim 0.019), confirming that contrastive
loss produces genuine weight-space differentiation. However, routing
selectivity is weak: both experts produce similar-magnitude deltas regardless
of which tokens are routed to them. This reveals \textbf{redundant representation}: the experts learned different
directions in representation space, but both directions are needed by most
tokens.

We initially hypothesized that this was caused by placing experts in a
domain-agnostic zone (layers 14--17), motivated by the layer divergence
analysis (Section~\ref{sec:divergence}). However, repeating the routing
analysis on the trunk-18 configuration (experts on layers 18--27, where CKA
$>$0.10) yielded the same result: both experts route 49--51\% of tokens
uniformly across all five domains, with per-expert PPL ablation varying by
only $\sim$0.26 points across domains. \textbf{Moving the fork to the
domain-divergent zone did not produce domain-specific routing.}

This suggests that the redundancy is not caused by fork location but by a more
fundamental limitation: contrastive loss differentiates expert \emph{weights}
but does not create functional \emph{routing specialization}. The router
learns to split tokens approximately 50/50 regardless of domain content,
and both experts contribute equally to all domains. Weight-space orthogonality
(CosSim $<$0.3) is a necessary but not sufficient condition for modularity.

% ============================================================================
\section{Related Work}
\label{sec:related}

\paragraph{LoRA and adapters.}
Hu et al.~\cite{hu2022lora} introduced Low-Rank Adaptation for efficient
fine-tuning. We extend LoRA from a fine-tuning tool to a \emph{structural unit
of specialization} within a single model. Li et al.~\cite{li2018intrinsic}
and Aghajanyan et al.~\cite{aghajanyan2021intrinsic} established that
fine-tuning operates on a low intrinsic dimension---our variable-rank mechanism
exploits this directly. HydraLoRA~\cite{hydralora2024} shares our core
insight: a shared A matrix captures functional patterns while multiple routed B
matrices handle domain-specific refinement. MoLoRA~\cite{molora2026} extends
per-token LoRA routing to achieve composable specialization, validating the
hot-swap vision independently.

\paragraph{Shared vs.\ routed experts.}
DeepSeekMoE~\cite{deepseek2024moe} introduced \emph{shared expert isolation}:
always-active experts absorb cross-domain knowledge, freeing routed experts to
specialize. DeepSeek-V3~\cite{deepseekv3} scales this to 1 shared + 256
routed experts. The ``Basic-Refinement Collaboration''
framework~\cite{basicrefinement2025} empirically confirms that shared experts
perform entity recognition and syntactic parsing while routed experts refine
domain-specific associations. Our finding that experts differentiate along
functional axes but not domain axes (Section~\ref{sec:magnitude}) is
independently reported by OpenMoE~\cite{openmoe2024} (context-independent
token-type routing), the Mixtral analysis~\cite{mixtral2024} (``no obvious
patterns based on topic''), and the Standing
Committee~\cite{standing2026} paper (functional experts as stable core, domain
experts as peripheral). The ``Domain vs.\ Driver Experts''
analysis~\cite{domaindriver2026} formalizes this: driver experts handle
function words and syntax, domain experts handle content vocabulary.

\paragraph{Specialization theory.}
B\'{e}na \& Goodman~\cite{bena2025modularity} prove resource constraints are
necessary for specialization---without cost pressure, modules converge. Our
contrastive loss provides one form of constraint; the rank measures the result.
However, our experiments show that weight-space constraints (contrastive loss)
produce representational but not functional specialization. The literature
suggests that top-$K$ routing ($K > 1$) may be required: with $K=1$, each
token must choose a single axis of specialization (typically functional), while
$K \geq 2$ allows simultaneous functional and domain expert
activation~\cite{standing2026, domaindriver2026}.

\paragraph{Low-rank routing.}
L2R~\cite{l2r2026} projects to $r{=}2$ dimensions for routing, confirming that
routing decisions are intrinsically low-dimensional. Our variable-rank adapters
extend this: the routing is low-rank, and so is the specialization itself.

% ============================================================================
\section{Understanding the Tree Structure}
\label{sec:understanding}

Beyond demonstrating that the framework works, we pose three scientific
questions about the emergent structure.

\paragraph{Q1: Does the rank distribution follow a power law?}
If specialization depth reflects the Zipf-distributed information density of
natural language~\cite{shannon1948}, we expect many low-rank lenses and few
high-rank crystals. We test this by fitting a power-law model to the rank
distribution across all experts at convergence.

\placeholder{Figure: Log-log plot of expert count vs.\ rank. If the
distribution is heavy-tailed, most experts are rank-1--4 lenses, with a long
tail of rank-16+ crystals. Report the power-law exponent and
goodness-of-fit.}

\paragraph{Q2: Does rank correlate with domain KL-divergence from the core?}
The unifying claim (Table~\ref{tab:unification}) predicts that an expert's rank
should correlate with how different its token distribution is from the
generalist core's. We measure this by computing $D_{\mathrm{KL}}(P_{\text{expert}}
\| P_{\text{core}})$ over the vocabulary distribution and plotting against
adapter rank.

\placeholder{Figure: Scatter plot of adapter rank vs.\
$D_{\mathrm{KL}}(\text{expert} \| \text{core})$. If the unification holds,
these should correlate positively. Report Spearman's $\rho$.}

\paragraph{Q3: Is the emergent tree structure interpretable?}
We examine whether the learned branching corresponds to human-recognizable
domain boundaries (e.g., science vs.\ narrative, formal vs.\ informal) or
whether it reflects statistical structure that cuts across human categories.

\placeholder{Figure: Visualization of the tree with representative tokens at
each leaf, colored by domain. Annotate whether branches align with identifiable
topics or cross-cut them.}

% ============================================================================
\section{Limitations}
\label{sec:limitations}

We identify four limitations that bound our current claims:

\begin{enumerate}
    \item \textbf{Weight differentiation does not imply functional
          specialization.} This is our most important negative finding.
          Contrastive loss reliably produces orthogonal expert weights
          (CosSim $<$0.3) at both fork boundaries tested (layers 14 and 18).
          However, token routing analysis shows both experts route all domains
          uniformly (49--51\% split), and removing either expert degrades all
          domains equally. The experts learn different \emph{directions} in
          representation space, but both directions are needed by all tokens.
          Achieving genuine functional modularity---where experts are
          independently removable with selective damage---remains an open
          problem.
    \item \textbf{No second fork triggered on generic data.} Despite testing
          bimodality thresholds from 0.3 to 0.555, no configuration produced
          more than 2 experts. On generic C4 data, 2-expert systems appear to
          be stable equilibria. Domain-specific curricula may be required to
          trigger deeper forking.
    \item \textbf{Framework tested on one base model only.} All experiments use
          Qwen3-1.7B. Generalization to other architectures (Llama, Mistral,
          GPT-scale) and other data distributions is assumed but not
          demonstrated.
    \item \textbf{Learntropy definition is provisional.} Our use of raw
          cross-entropy as a splitting signal does not distinguish productively
          difficult tokens from noise (Section~\ref{sec:method}). A principled
          learntropy function that models net learning value---closer to
          Wozniak's~\cite{wozniak2018learntropy} definition---may be needed
          for effective curriculum-driven forking.
\end{enumerate}

% ============================================================================
\section{Conclusion}

We set out to test whether specialization depth, adapter rank, and storage tier
are the same dimension. The evidence is mixed.

\textbf{What works:} Contrastive LoRA forking reliably produces weight-space
differentiation (CosSim $<$0.3) where six prior approaches achieved 0.998 at
best. Layer divergence analysis identifies where domain signals emerge (layer
18 in Qwen3-1.7B), and placing experts in this zone accelerates
differentiation. Variable-rank adapters measured in kilobytes can improve
perplexity from 21.20 to 19.87 with only 655K trainable parameters.

\textbf{What does not work yet:} Weight-space differentiation does not produce
functional modularity. Experts route tokens uniformly across domains and
cannot be independently removed without uniform quality degradation. No
configuration triggered a second fork on generic C4 data. The tree structure
we hypothesized has not emerged.

\textbf{What this means:} The \tok{} framework demonstrates that LoRA
adapters can serve as a structural unit of specialization---compact,
variable-rank, and trainable on commodity hardware. But achieving the
modular, hot-loadable expert tree requires mechanisms beyond contrastive loss:
possibly domain-aware curricula, goal-directed routing~\cite{wozniak2018learntropy},
or explicit functional objectives that reward routing selectivity rather than
weight orthogonality.

% ============================================================================
\appendix

\section{Mathematical Details}
\label{app:math}

\subsection{Learntropy Bimodality Test}

Given an expert $e$ processing tokens $\{x_1, \ldots, x_T\}$, we compute the
per-token loss vector $\ell_e = [\ell_e(x_1), \ldots, \ell_e(x_T)]$ and apply
Hartigan's dip test~\cite{hartigan1985dip} for unimodality. The expert is a
split candidate when the dip statistic exceeds a threshold $\delta$:
\begin{equation}
    \text{DIP}(\ell_e) > \delta \implies \text{split expert } e
\end{equation}
We set $\delta = 0.05$ (the conventional significance level for rejecting
unimodality).

\subsection{Rank Growth Criterion}

After a split, each child adapter begins at rank $r = 1$. Let
$W_{\text{parent}}$ be the parent's weight matrix and $\Delta W = BA$ the LoRA
perturbation at current rank $r$. We compute the relative reconstruction error:
\begin{equation}
    \epsilon_r = \frac{\|W_{\text{target}} - W_{\text{parent}} - BA\|_F}{\|W_{\text{target}} - W_{\text{parent}}\|_F}
\end{equation}
where $W_{\text{target}}$ is the optimal weight matrix estimated from training
data. When $\epsilon_r > \epsilon_{\max}$, the rank is doubled: $r \leftarrow 2r$.

\subsection{Combined Training Objective}

The full training loss for each expert combines the standard language modeling
loss with the contrastive term:
\begin{equation}
    \mathcal{L} = \mathcal{L}_{\text{LM}} + \lambda_c \sum_{i < j} \max\bigl(0,\; \cos(\theta_{ij}) - \tau\bigr)
\end{equation}
We use $\lambda_c = 0.1$ and $\tau = 0.85$ in all experiments.

\section{Shannon's Entropy Argument for Zipf-Aligned Routing}
\label{app:shannon}

Shannon's analysis of English~\cite{shannon1948} established that natural
language has a specific entropy structure: highly redundant on average
($\sim$1 bit per character), but with information concentrated in sparse,
high-surprise moments. This is the source of Zipf's law---common tokens carry
little information and appear frequently, while rare tokens carry much
information and appear rarely.

Standard MoE load-balancing ignores this structure: it allocates equal capacity
to every expert regardless of information content. \tok{}'s argument is a
Shannon argument: parameter allocation should reflect information density. When
specialization depth follows Zipf's law, so does adapter rank, and therefore so
does storage footprint. The most frequently needed adapters are the smallest;
the largest adapters are needed least often. This access pattern is precisely
what tiered storage hierarchies exploit---not by coincidence, but because both
language and memory hierarchies are shaped by power-law economics.

At current scale (Qwen3-1.7B, adapters totaling $<$20\,MB), tiered storage is
unnecessary---everything fits in VRAM. Tiered deployment becomes relevant at
three thresholds: (1)~larger cores (70B+) where VRAM is contested,
(2)~thousands of community-contributed experts, and (3)~edge devices with
6--8\,GB total memory.

\section{Tiered Storage Validation}
\label{app:tiered}

We conducted 121 inference experiments on rented GPU instances (NVIDIA RTX PRO
6000 Blackwell, OLMoE-1B-7B) to calibrate a hardware-realistic cache simulator.

\begin{table}[h]
\centering
\caption{Memory tier economics (Netherlands retail, March 2026, incl.\ VAT).}
\label{tab:tiers}
\begin{tabular}{@{}lrrr@{}}
\toprule
Medium & Capacity & Bandwidth & Cost/GB \\
\midrule
GPU VRAM (GDDR7) & 48\,GB & $\sim$1.8\,TB/s & (fixed at mfg.) \\
DDR5-5600 & 256\,GB & $\sim$90\,GB/s & $\sim$\texteuro{}12/GB \\
NVMe Gen5 & 2--4\,TB & $\sim$14\,GB/s & $\sim$\texteuro{}0.09/GB \\
NVMe Gen4 & 2--4\,TB & $\sim$7\,GB/s & $\sim$\texteuro{}0.07/GB \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[h]
\centering
\caption{Expert load time from Gen5 NVMe at 14\,GB/s, vs.\ per-token compute
time at 134\,tok/s (7.5\,ms/token).}
\label{tab:load-times}
\begin{tabular}{@{}lrrl@{}}
\toprule
Adapter rank & Size & Load time & Critical path? \\
\midrule
Rank-1 & 4\,KB & 0.3\,$\mu$s & No (CPU cache) \\
Rank-16 & 64\,KB & 4.6\,$\mu$s & No ($<$0.1\% of token time) \\
Rank-64 & 256\,KB & 18\,$\mu$s & No ($<$0.3\% of token time) \\
Full expert (OLMoE) & 16\,MB & 1.1\,ms & Marginal (15\% of token time) \\
\bottomrule
\end{tabular}
\end{table}

Even at rank-64, adapter loads consume $<$0.3\% of a token's compute budget.
Tiered storage is not a compromise---it is invisible at \tok{} adapter sizes.

% ============================================================================
\begin{thebibliography}{99}
\bibitem{hu2022lora} E. Hu et al. LoRA: Low-Rank Adaptation of Large Language Models. \textit{ICLR}, 2022.
\bibitem{geva2021transformer} M. Geva, R. Schuster, J. Berant, O. Levy. Transformer Feed-Forward Layers Are Key-Value Memories. \textit{EMNLP}, 2021.
\bibitem{li2018intrinsic} C. Li et al. Measuring the Intrinsic Dimension of Objective Landscapes. \textit{ICLR}, 2018.
\bibitem{aghajanyan2021intrinsic} A. Aghajanyan et al. Intrinsic Dimensionality Explains the Effectiveness of Language Model Fine-Tuning. \textit{ACL}, 2021.
\bibitem{komatsuzaki2023sparse} A. Komatsuzaki et al. Sparse Upcycling. \textit{arXiv:2212.05055}, 2023.
\bibitem{dropupcycle2025} Drop-Upcycling: Training Sparse MoE with Partial Re-initialization. \textit{ICLR}, 2025.
\bibitem{roller2021hash} S. Roller et al. Hash Layers For Large Sparse Models. \textit{NeurIPS}, 2021.
\bibitem{emoe2026} EMoE: Eigenbasis-Guided Routing. \textit{arXiv:2601.12137}, 2026.
\bibitem{idamoe2025} IDA-MoE: Decoupled Routing via GMMs. \textit{ACM MM}, 2025.
\bibitem{bena2025modularity} G. B\'{e}na, D. Goodman. Dynamics of Specialization Under Resource Constraints. \textit{Nature Communications}, 2025.
\bibitem{l2r2026} L2R: Low-Rank Lipschitz-Controlled Routing. \textit{arXiv:2601.21349}, 2026.
\bibitem{shannon1948} C. E. Shannon. A Mathematical Theory of Communication. \textit{Bell System Technical Journal}, 27(3):379--423, 1948.
\bibitem{hartigan1985dip} J. A. Hartigan, P. M. Hartigan. The Dip Test of Unimodality. \textit{Annals of Statistics}, 13(1):70--84, 1985.
\bibitem{piaget1952} J. Piaget. \textit{The Origins of Intelligence in Children}. International Universities Press, 1952.
\bibitem{vygotsky1978} L. S. Vygotsky. \textit{Mind in Society: The Development of Higher Psychological Processes}. Harvard University Press, 1978.
\bibitem{wozniak2018learntropy} P. A. Wozniak. Learntropy. \textit{SuperMemo Guru}, 2018. \url{https://supermemo.guru/wiki/Learntropy}.
\bibitem{standing2026} The Illusion of Specialization: Standing Committee in Mixture-of-Experts. \textit{arXiv:2601.03425}, 2026.
\bibitem{deepseek2024moe} X. Dai et al. DeepSeekMoE: Towards Ultimate Expert Specialization. \textit{ACL}, 2024.
\bibitem{deepseekv3} DeepSeek-AI. DeepSeek-V3 Technical Report. \textit{arXiv:2412.19437}, 2024.
\bibitem{hydralora2024} T. Tian et al. HydraLoRA: An Asymmetric LoRA Architecture for Efficient Fine-Tuning. \textit{NeurIPS}, 2024.
\bibitem{molora2026} MoLoRA: Composable Specialization via Per-Token Adapter Routing. \textit{arXiv:2603.15965}, 2026.
\bibitem{basicrefinement2025} Decoding Knowledge Attribution in MoE: A Framework of Basic-Refinement Collaboration. \textit{ACL}, 2025.
\bibitem{openmoe2024} F. Xue et al. OpenMoE: An Early Effort on Open Mixture-of-Experts Language Models. \textit{ICML}, 2024.
\bibitem{mixtral2024} A. Jiang et al. Mixtral of Experts. \textit{arXiv:2401.04088}, 2024.
\bibitem{domaindriver2026} What Gets Activated: Uncovering Domain and Driver Experts in MoE Language Models. \textit{arXiv:2601.10159}, 2026.
\end{thebibliography}

\end{document}
