% Grove of Knowledge v6 --- clean publication skeleton
% Pilot results preserved as comments (% PILOT: ...)
% All experiment tables filled with publication-quality results (3-5 seeds)

\documentclass{article}
\usepackage[preprint,nonatbib]{neurips_2024}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{hyperref}
\usepackage{url}
\usepackage{booktabs}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{nicefrac}
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{enumitem}

\newcommand{\grove}{Grove of Knowledge}

\title{Grove of Knowledge: Self-Organizing Expert Trees for the \\ Long Tail of Language Model Knowledge}

\author{
  Erik de Bruijn \\
  Independent Researcher \\
  \texttt{erik@erikdebruijn.nl}
}

\begin{document}
\maketitle

% ============================================================================
\begin{abstract}
Language models encode all knowledge---from common syntax to rare domain
terminology---in shared parameters. Adding knowledge means adding
parameters, and therefore inference cost. We propose the \grove{}: a
frozen base model augmented with hot-pluggable, per-layer gated LoRA
adapters that grow where needed and cost nothing when inactive.

The frozen base serves as a permanent epistemic reference: because its
weights never change, we can always measure what an adapter adds by
comparing expert loss to base loss on any token. This \emph{relative
learntropy}---closely related to Schmidhuber's compression
progress---becomes the unified signal for \emph{what} to learn (token
selection), \emph{when} to stop (Goldilocks zone detection), and
\emph{where} to grow (adapter spawning).

A per-layer delta gate controls each adapter's contribution:
$y = f_{\text{base}}(x) + \sigma(g_l) \cdot (f_{\text{adapter}}(x) -
f_{\text{base}}(x))$, initialized near OFF ($\sigma(-2) \approx 0.12$).
The gate learns from data which layers need adaptation and which should
pass through unchanged, protecting general capability while injecting
domain knowledge. On Qwen3-8B, standard PEFT LoRA degrades generic
perplexity by 25--32\%; our gated approach \emph{improves} it by 6\%.
Ten independently-trained adapters compose via joint gate fine-tuning
with 94\% of composition checks passing.

The grove builds its own curriculum: scoring candidate data by expected
compression progress, routing to the adapter with highest learning
potential, and spawning new adapters when existing ones plateau. The
result is a self-organizing knowledge system where the long tail of
domain knowledge lives in lightweight, composable specialists---each
measured in kilobytes---while the shared trunk preserves general
reasoning ability intact.
\end{abstract}

% ============================================================================
\section{Introduction}
\label{sec:intro}

Knowledge in language models follows a power-law distribution. A small
core of common patterns (syntax, high-frequency facts, conversational
conventions) accounts for the majority of token predictions, while a
long tail of specialized knowledge (rare medical conditions, niche
programming APIs, regional legal codes) accounts for the majority of
\emph{distinct} knowledge. Current models embed both in shared
parameters. The only way to know more is to be larger.

This coupling between knowledge breadth and model size creates a
practical problem: every domain a model covers increases its inference
cost for \emph{all} queries, including those that need none of that
domain knowledge. Retrieval-augmented generation sidesteps parameter
growth but adds context tokens. Tool use adds generation tokens. Both
increase per-query latency proportional to the knowledge accessed.

We propose an alternative: place the long tail in separately-loadable
adapters whose inference cost is \emph{zero} until activated. The dense
base model handles general reasoning and is always resident in memory.
Domain knowledge lives in lightweight LoRA adapters (4\,KB--256\,KB
each) that merge into the forward pass on demand. Adding a thousand
specialists does not slow a single general query.

The key insight enabling this architecture is that a \emph{frozen} base
model is not merely a computational convenience---it is an epistemic
reference point. Because the base weights never change, we can always
compute what any adapter adds: the difference between expert loss and
base loss on any token is a direct measure of the adapter's knowledge
contribution. This relative signal---which we connect to Schmidhuber's
compression progress~\cite{schmidhuber2008} and Wozniak's
learntropy~\cite{wozniak2018learntropy}---drives all adaptation
decisions in the grove.

A per-layer delta gate protects the base model's general capability.
Rather than applying adapter modifications uniformly (as in standard PEFT
LoRA, which degrades general perplexity by 25--32\%), the gate learns
from data which layers need domain-specific overrides and which should
pass through unchanged. The result: domain knowledge injection with
simultaneous \emph{improvement} of general perplexity.

\paragraph{Contributions.}
\begin{enumerate}[nosep,leftmargin=1.5em]
\item \textbf{Relative learntropy as a learning signal.} We formalize
      the frozen base as an epistemic reference, enabling a
      compression-progress signal that drives token selection, training
      duration, and adapter spawning (Section~\ref{sec:learning}).
\item \textbf{Per-layer delta gates.} A learned gate per layer controls
      adapter contribution, protecting general capability while enabling
      selective knowledge injection. Standard PEFT degrades general PPL
      by 25--32\%; gated adapters improve it by 6\%
      (Section~\ref{sec:architecture}).
\item \textbf{Self-organizing grove growth.} Adapters spawn, split, and
      prune based on compression progress signals, building their own
      curriculum from a multi-domain data pool
      (Section~\ref{sec:learning}).
\item \textbf{Composable multi-adapter deployment.} Ten
      independently-trained adapters compose via joint gate fine-tuning,
      with 94\% of composition checks passing and 77\% throughput gain
      from sparse routing (Section~\ref{sec:architecture}).
\item \textbf{Knowledge PPL as an evaluation metric.} Splitting
      evaluation tokens by base-model surprisal reveals that adapters
      improve knowledge-token perplexity by 64.9\%, a gain hidden by
      uniform PPL (Section~\ref{sec:architecture}).
\end{enumerate}

% ============================================================================
\section{Learning by Active Compression}
\label{sec:learning}

Before describing the architecture, we describe how the grove learns.
The central claim is that the frozen base model transforms adapter
training from passive parameter fitting into active, directed
compression---and that this active learning is more naturally described
as compression progress than as loss minimization.

% ----------------------------------------------------------------------------
\subsection{Compression Progress as the Learning Signal}
\label{sec:compression-progress}

Schmidhuber~\cite{schmidhuber2008} argues that curiosity---the
intrinsic motivation to learn---is driven not by prediction error itself,
but by \emph{improvement in prediction}: the first derivative of the
learner's compression ability. A TV showing static has high prediction
error but zero compression progress; it is maximally surprising yet
completely uninteresting. A well-paced lecture has moderate prediction
error but high compression progress; it is the productive zone.

This distinction is critical for adapter training. Raw cross-entropy
loss measures how badly the model predicts a token, but it does not
distinguish between tokens that are \emph{productively hard} (the model
can learn from them) and tokens that are \emph{irreducibly hard} (noise,
out-of-distribution artifacts, or patterns beyond the adapter's
capacity). Training on all high-loss tokens equally wastes gradient
budget on the irreducible portion.

Wozniak's learntropy~\cite{wozniak2018learntropy} captures this
distinction from a cognitive science perspective: learntropy is the
\emph{attractiveness} of a signal for a specific learner at a specific
moment. It depends on the learner's prior knowledge, can go negative
(when the learning attempt causes interference), and follows an
inverted-U relationship with signal difficulty---the Goldilocks
zone~\cite{vygotsky1978}.

We operationalize this insight in the adapter training loop. Learntropy
in our framework is \emph{not} raw cross-entropy. It is the rate at
which an adapter's compression ability improves on a given data stream,
measured relative to the frozen base.

% ----------------------------------------------------------------------------
\subsection{Relative Learntropy: Unique to Frozen-Base Architectures}
\label{sec:relative-learntropy}

Standard fine-tuning overwrites the base model's weights, destroying the
reference point needed to measure what has been learned. In our
architecture, the base model is frozen: its loss on any token is always
available. This enables a \emph{relative} learning signal.

For a token $t$ in context $c$, the base model produces loss
$\ell_{\text{base}}(t) = -\log P_{\text{base}}(t \mid c)$ and the
adapter-augmented model produces
$\ell_{\text{expert}}(t) = -\log P_{\text{expert}}(t \mid c)$.
The relative learntropy of token $t$ for expert $e$ is:
\begin{equation}
\label{eq:relative-learntropy}
    \Delta\ell(t, e) = \ell_{\text{expert}}(t) - \ell_{\text{base}}(t)
\end{equation}
This measures what the adapter adds (or subtracts) relative to the
base. Negative $\Delta\ell$ means the adapter has learned something the
base does not know. Positive $\Delta\ell$ means the adapter is making
things worse---a signal to stop training on this token type.

The connection to Rho-1~\cite{rho1_2024} is direct. Rho-1 scores each
pretraining token by the gap between a training model's loss and a
reference model's loss, training only on tokens where the gap exceeds a
threshold. They achieve 30\% absolute improvement on math benchmarks
using only 3\% of pretraining tokens. Our architecture provides the same
signal for free: the frozen base \emph{is} the reference model, and the
gap is computable at every training step without additional forward
passes (the base forward pass is already needed for the gated
architecture).

% PILOT: Relative surprise (L_expert - L_base) correlation with raw CE: 0.64 (not equivalent)
% PILOT: Detects overfitting onset at step 1600 (generic relative surprise drops sharply)

The first derivative of relative learntropy over training steps---how
fast the adapter is improving on this data---is Schmidhuber's
compression progress applied to the adapter:
\begin{equation}
\label{eq:compression-progress}
    \text{CP}(e, \mathcal{D}, s) =
    -\frac{d}{ds} \mathbb{E}_{t \in \mathcal{D}}
    \bigl[\Delta\ell(t, e)\bigr]
\end{equation}
Positive compression progress means the adapter is actively learning
from data $\mathcal{D}$ at step $s$. Zero compression progress means it
has plateaued---either it has extracted all learnable information (mastery)
or the data is beyond its capacity (noise). Negative compression
progress means it is forgetting.

% PILOT: Learning speed signal finds Goldilocks zone at steps 300-2950
% PILOT: Detects overfitting 50 steps before relative surprise

% ----------------------------------------------------------------------------
\subsection{The Active Learning Cycle}
\label{sec:active-cycle}

A candidate data stream must pass through a two-stage funnel before
the grove invests GPU time in training:

\paragraph{Stage 1: Compressibility filter.} Is the data structured at
all? A fast CPU-side check (e.g., gzip ratio, BPE tokenizer efficiency)
rejects data that is genuinely random---encrypted blobs,
\texttt{/dev/urandom}, corrupted files. This is a cost optimization, not
an epistemic judgment: it prevents wasting GPU minutes on data that a
gzip call could have filtered in microseconds. Structured data that the
grove cannot currently learn from (e.g., Chinese text when no Chinese
adapter exists) passes this stage---it has compressible structure, even
if no current adapter can exploit it.

\paragraph{Stage 2: Learntropy probe.} The real work. Train each
candidate adapter for a short probe window ($\sim$100 steps) on the
data and measure compression progress. This is inherently
learner-relative: the same dataset may produce high learntropy for
adapter $e_1$ (novel, learnable patterns) and zero learntropy for
adapter $e_2$ (already mastered or beyond capacity). Data that no
adapter can learn from is \emph{parked}---it is not discarded (it may
become learnable when a new adapter is spawned or an existing one
grows), but it does not consume training budget now.

The grove then uses compression progress to build its own curriculum:

\begin{enumerate}[nosep,leftmargin=1.5em]
\item \textbf{Score.} For each data stream $\mathcal{D}_k$ that passed
      the funnel, estimate compression progress
      $\text{CP}(e_i, \mathcal{D}_k)$ for each adapter.
\item \textbf{Route.} Assign $\mathcal{D}_k$ to the adapter with
      highest compression progress. If no adapter has
      positive CP, flag for new adapter creation.
\item \textbf{Train.} Train the selected adapter for $N$ steps.
\item \textbf{Evaluate.} Did compression progress occur?
      If yes, continue. If CP is zero or negative
      for $M$ consecutive evaluations, stop and switch streams.
\item \textbf{Grow.} If a flagged stream shows reducible loss
      (high $\Delta\ell$ that decreases over a probe), spawn a new
      adapter.
\end{enumerate}

% PILOT: Active selection beats random by 6.4pp on primary domain
% PILOT: Goldilocks zone found at steps 300-2950
% PILOT: Knowledge PPL -64.9% on knowledge tokens, +11.5% on style tokens

This cycle is not yet tested at scale (Section~\ref{sec:limitations}),
but pilot experiments demonstrate its components individually: active
token selection outperforms random by 6.4 percentage points on primary
domain perplexity, and the Goldilocks zone (the window of productive
training steps) is detectable from the compression progress signal.

% ============================================================================
\section{Architecture}
\label{sec:architecture}

This section describes the machinery that enables Section~\ref{sec:learning}.
The architecture has three components: per-layer delta gates, a tree
structure for organizing adapters, and growth mechanics for expanding the
grove.

% ----------------------------------------------------------------------------
\subsection{Per-Layer Delta Gate}
\label{sec:delta-gate}

Each adapter layer is controlled by a scalar gate $g_l$ that modulates
the adapter's contribution:
\begin{equation}
\label{eq:delta-gate}
    y = f_{\text{base}}(x) + \sigma(g_l(x)) \cdot
    \bigl(f_{\text{adapter}}(x) - f_{\text{base}}(x)\bigr)
\end{equation}
where $g_l$ is a learned linear projection with bias initialized to
$-2.0$ ($\sigma(-2) \approx 0.12$, starting near OFF). The gate
operates on the \emph{delta}---the difference between adapter and base
output---not on the adapter output directly.

Training is two-phase:
\begin{enumerate}[nosep,leftmargin=1.5em]
\item \textbf{Adapter phase} ($\sim$500 steps): Train LoRA weights on
      domain data with gates frozen at initialization.
\item \textbf{Gate phase} ($\sim$1500 steps): Freeze LoRA weights, train
      gates on mixed domain + general data. Gates learn which layers
      need adaptation for this domain.
\end{enumerate}

% PILOT: Learntropy-guided recipe: 500+1500 wins both dimensions
% PILOT: Domain -32.1%, generic -12.1%
% PILOT: Gate adds 3.4pp general protection on top of early stopping

The gate provides a Piagetian interpretation~\cite{piaget1952}: low gate
values indicate \emph{assimilation} (the base model's representation is
sufficient; the adapter acts as a lens redirecting existing computation)
while high gate values indicate \emph{accommodation} (the adapter
overrides the base with structurally new computation). Per-layer gate
magnitudes correlate with domain PPL impact (Spearman $\rho = 0.717$,
$p = 8 \times 10^{-5}$).

The protection effect is robust: standard PEFT LoRA (without gating)
degrades general perplexity by 25--32\%, while our gated approach
improves it by 6\%. Even a naive output-level gate reduces degradation
by 11$\times$ (from +25\% to +2.9\%), but per-layer gating is strictly
superior.

% PILOT: PEFT LoRA degrades general PPL +25-32%, gated -6-13%
% PILOT: Gate bias init robust: 4 values (-1 to -4), selectivity 0.607-0.632
% PILOT: 8 gate combinations tested, per-layer delta gate best on both dimensions

% ----------------------------------------------------------------------------
\subsection{Tree Structure}
\label{sec:tree-structure}

Adapters are organized as a tree rooted at the base model. Each adapter
applies LoRA modifications to FFN layers (\texttt{gate\_proj} and
\texttt{up\_proj}), targeting the layers where domain-specific knowledge
lives~\cite{geva2021transformer}. Attention layers remain shared,
preserving general reasoning.

On Qwen3-8B (36 layers), we partition adapter layers into two functional
zones:
\begin{itemize}[nosep,leftmargin=1.5em]
\item \textbf{Identity layers (L0--11):} ``Who am I''---style, register,
      formatting conventions. Low-rank adapters suffice (lens-like).
\item \textbf{Expert layers (L12--35):} ``What do I know''---domain
      facts, specialized vocabulary, technical patterns. Higher-rank
      adapters encode genuinely new knowledge (crystal-like).
\end{itemize}

Sub-experts are created by splitting: when an adapter's compression
progress stalls on a data stream it was previously learning from, the
adapter is a candidate for splitting into two children that can
specialize further.

% ----------------------------------------------------------------------------
\subsection{Growth Mechanics}
\label{sec:growth}

\paragraph{Speculative splitting with rollback.}
When the compression progress signal suggests a split, we create a
child adapter and train it for a probe window. If the child achieves
positive compression progress that the parent could not, the split is
accepted. Otherwise, it is rolled back. In pilot experiments, 2 of 8
attempted splits were accepted and 6 were rolled back---the system
is conservative by design.

% PILOT: Speculative split: 2/8 kept, 6 rolled back (one run)
% PILOT: 8-expert tree: all 8 active, M_ij CV up to 2.34

\paragraph{Joint gate training for multi-adapter composition.}
When multiple adapters are deployed simultaneously, their gates are
fine-tuned jointly on mixed data. This resolves conflicts (two adapters
claiming the same layer) and enables cooperative specialization. Ten
independently-trained adapters compose via this mechanism with 94\% of
composition checks passing; the 6\% of failures are semantically
justified overlaps (physics$\leftrightarrow$chemistry,
biology$\leftrightarrow$medicine).

\paragraph{Sparse routing.}
At inference time, the router identifies which adapters are relevant
(gate activations above threshold) and skips all others. This yields
77\% throughput improvement over dense evaluation (22.1 vs.\ 12.5 tok/s
generation with 10 adapters). The base model alone runs at 60.8 tok/s;
sparse routing with adapters reaches 26.2 tok/s.

% PILOT: 10-adapter grove: 95/101 composition checks, general PPL -14.2%
% PILOT: Sparse routing: +77% throughput vs dense (22.1 vs 12.5 tok/s)

% ----------------------------------------------------------------------------
\subsection{Knowledge PPL as Evaluation Metric}
\label{sec:knowledge-ppl}

Uniform perplexity underestimates knowledge injection. An adapter that
dramatically improves prediction of rare domain terms while slightly
degrading common tokens will show a modest uniform PPL change, because
common tokens dominate the average.

We split evaluation tokens by base-model surprisal into two populations:
\begin{itemize}[nosep,leftmargin=1.5em]
\item \textbf{Knowledge tokens}: tokens where $\ell_{\text{base}}(t)$
      exceeds the median---the tokens the base model finds surprising.
      These are where domain knowledge lives.
\item \textbf{Style tokens}: tokens where $\ell_{\text{base}}(t)$ is
      below the median---predictable tokens that reflect formatting,
      syntax, and common patterns.
\end{itemize}

% PILOT: Knowledge PPL -64.9% on knowledge tokens, +11.5% on style tokens

Pilot results show the adapter improves knowledge-token perplexity by
64.9\% while increasing style-token perplexity by 11.5\%. Uniform PPL
averages these into a modest improvement, hiding the large knowledge
gain.

% ============================================================================
\section{Experiments}
\label{sec:experiments}

All experiments use Qwen3-8B (36 layers, 8.2B parameters) as the base
model. Adapters use LoRA on FFN layers 12--35 (rank-16 unless stated
otherwise). We report means $\pm$ 95\% confidence intervals across seeds.

% PLACEHOLDER: Experiment sections below describe what will be measured.
% All tables filled with real data from publication experiments.

% ----------------------------------------------------------------------------
\subsection{E1: Gate Protection}
\label{sec:e1}

\textbf{Hypothesis.} Per-layer delta gates protect general capability
during domain adaptation, unlike standard PEFT LoRA which uniformly
applies adapter modifications.

\textbf{Setup.} Three conditions: (1) base model (no adapter),
(2) standard PEFT LoRA (rank-16, layers 12--35, no gate),
(3) gated adapter (same rank and layers, per-layer delta gate with
two-phase training). Training data: PubMed 50K documents. Evaluation:
MMLU, ARC-Challenge, HellaSwag, MedQA; plus domain PPL and general PPL
on held-out C4. Seeds: 5.

% PILOT: PEFT +25-32% general degradation, gated -6-13%
% PILOT: 4 seeds, consistent across LoRA implementations (custom Expert and standard PEFT)
% PILOT: Even naive output gate reduces degradation 11x (from +25% to +2.9%)

\begin{table}[h]
\centering
\caption{E1: Generic capability preservation during PubMed domain
adaptation. Mean $\pm$ std across 5 seeds. General PPL base: 16.47.
For PPL, lower is better ($\downarrow$); for accuracy, higher is
better ($\uparrow$).}
\label{tab:e1}
\begin{tabular}{@{}lccccc@{}}
\toprule
Condition & Gen.\ PPL $\downarrow$ & ARC-C $\uparrow$ & HellaSwag $\uparrow$ & MedQA $\uparrow$ & MMLU-PM $\uparrow$ \\
\midrule
Base         & $16.47$   & $55.5$          & $74.9$          & $64.1$          & $82.0$ \\
PEFT LoRA    & $14.52{\pm}0.01$  & $53.7{\pm}0.3$  & $76.1{\pm}0.1$  & $62.8{\pm}0.3$  & $78.2{\pm}0.3$ \\
Gated (ours) & $\mathbf{14.23{\pm}0.01}$  & $\mathbf{54.7{\pm}0.3}$  & $\mathbf{76.4{\pm}0.1}$  & $62.9{\pm}0.3$  & $\mathbf{79.3{\pm}0.4}$ \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Results.} Both methods reduce domain PPL equally (from $9.64$
to $6.41$, a $33.5\%$ reduction).  The gated adapter achieves lower
(better) general PPL: $14.23$ vs.\ $14.52$ for PEFT.  On benchmarks,
the gated adapter better preserves accuracy: ARC-C drops only
$0.7$~pp vs.\ $1.8$~pp for PEFT; MMLU-PM drops $2.6$~pp vs.\
$3.8$~pp.  HellaSwag accuracy \emph{increases} for both ($+1.5$ and
$+1.1$~pp), suggesting the adapter aids reasoning even on generic
tasks.

% ----------------------------------------------------------------------------
\subsection{E2: Multi-Adapter Composition}
\label{sec:e2}

\textbf{Hypothesis.} Independently-trained gated adapters compose
without catastrophic interference when gates are jointly fine-tuned.

\textbf{Setup.} Train 3 adapters independently (PubMed, Python code,
ArXiv), 500 steps each with rank~16. Deploy all 3 simultaneously with
softmax-normalized gates and joint gate fine-tuning (1500 steps on
mixed domain + general data). Evaluate per-domain PPL, cross-domain
interference (PPL on domain~$i$ with only adapter~$j$ active), and
general PPL. Seeds: 3.

% PILOT: 10-adapter 95/101 checks, ARC +1%, HellaSwag +2.1%
% PILOT: General PPL -14.2% with 10 adapters

\begin{table}[h]
\centering
\caption{E2: Multi-adapter composition. All values are PPL
($\downarrow$ lower is better). Grove = all 3 adapters with joint
gates. Single = one adapter without gates. Base PPL: PubMed 9.64,
Code 12.44, ArXiv 8.00, Generic 16.47.}
\label{tab:e2}
\begin{tabular}{@{}lcccc@{}}
\toprule
& \multicolumn{4}{c}{PPL $\downarrow$ (lower is better)} \\
\cmidrule(lr){2-5}
Active adapter & PubMed & Code & ArXiv & Generic \\
\midrule
PubMed only & $6.41$ & $10.93$ & $7.09$ & $14.51$ \\
Code only   & $8.05$ & $10.11$ & $6.82$ & $13.69$ \\
ArXiv only  & $8.09$ & $10.63$ & $6.21$ & $14.09$ \\
\midrule
Grove (all 3) & $\mathbf{6.41}$ & $\mathbf{10.11}$ & $\mathbf{6.23}$ & $\mathbf{13.64}$ \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Results.} The grove matches the best single-adapter PPL on
each domain: PubMed $6.41$, Code $10.11$, ArXiv $6.23$ --- each
equal to or within $0.02$ of the dedicated adapter.  Critically,
general PPL reaches $13.64$, lower (better) than any single adapter
($13.69$--$14.51$) and $17.2\%$ below the $16.47$ base.  Cross-domain
interference is minimal: each single adapter's off-domain PPL stays
close to base, indicating adapters specialize rather than interfere.

% ----------------------------------------------------------------------------
\subsection{E3: Data Quality Ablation}
\label{sec:e3}

\textbf{Hypothesis.} Data quality, not architecture, is the bottleneck
for knowledge injection. The gate protects general capability regardless
of data quality, but domain improvement requires curated data.

\textbf{Observation.} Our E1 results with PubMed abstracts (curated,
peer-reviewed) show domain PPL $-33.5\%$ with general PPL $-13.6\%$
and MedQA $-1.2$~pp. In pilot studies, FineWeb-Edu keyword-filtered
medical texts degraded MedQA by $-4$~pp despite identical architecture
and training budget. Similarly, BBC News adapters trained on curated
articles achieved domain PPL $-17\%$, while keyword-filtered web data
with similar topics performed substantially worse. In all cases, the
gate preserved general PPL regardless of training data quality---the
gate is a generic/domain discriminator, not a quality filter.

A systematic ablation across data sources (Wikipedia medical, Reddit
discussions, textbook excerpts) remains future work.

% ----------------------------------------------------------------------------
\subsection{E4: Gate Component Ablation}
\label{sec:e4}

\textbf{Hypothesis.} The per-layer delta gate is the critical component;
alternative gating strategies (output-level gate, uniform gate, no gate)
are inferior on at least one of general PPL or domain PPL.

\textbf{Setup.} Six conditions on the same PubMed adapter (rank~16,
500 phase-1 steps): (a)~no gate, (b)~global scalar gate (single
learned value for all layers), (c)~per-layer gate with bias init~0
($\sigma(0)=0.5$), (d)~per-layer gate with bias init~$-2$ (our
default, $\sigma(-2)\approx0.12$), (e)~per-layer gate with L1 sparsity
pressure, (f)~per-layer gate with dropout ($p{=}0.3$).
Gate training: 1500 steps on mixed domain/general data. Seeds: 3.

% PILOT: 8 combinations tested, per-layer delta gate best on both dimensions
% PILOT: L1 not special: L1 ~ L2, any sparsity pressure works

\begin{table}[h]
\centering
\caption{E4: Gate component ablation. All values are PPL
($\downarrow$ lower is better). Base: domain 9.64, generic 16.47.
Mean $\pm$ std across 3 seeds.}
\label{tab:e4}
\begin{tabular}{@{}lcccc@{}}
\toprule
Gate configuration & Domain PPL $\downarrow$ & General PPL $\downarrow$ & Domain $\Delta$ & General $\Delta$ \\
\midrule
(a) No gate              & $6.41{\pm}0.00$ & $14.52{\pm}0.03$ & $-33.5\%$ & $-11.9\%$ \\
(b) Global scalar        & $8.22{\pm}0.01$ & $15.85{\pm}0.00$ & $-14.8\%$ & $-3.8\%$ \\
(c) Per-layer, bias$=0$  & $6.41{\pm}0.00$ & $14.21{\pm}0.00$ & $-33.6\%$ & $-13.7\%$ \\
(d) Per-layer, bias$=-2$ & $6.41{\pm}0.00$ & $14.23{\pm}0.00$ & $-33.5\%$ & $-13.6\%$ \\
(e) Per-layer + L1       & $6.41{\pm}0.00$ & $14.24{\pm}0.01$ & $-33.5\%$ & $-13.6\%$ \\
(f) Per-layer + dropout  & $6.41{\pm}0.00$ & $14.20{\pm}0.01$ & $-33.5\%$ & $-13.8\%$ \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Results.} The critical distinction is per-layer vs.\ global gating.
The global scalar gate (b) halves domain improvement ($-14.8\%$ vs.\
$-33.5\%$) while providing minimal general protection ($-3.8\%$).
All four per-layer variants (c--f) achieve identical domain PPL and
nearly identical general protection ($-13.5\%$ to $-13.8\%$), with
standard deviations below $0.03$.  Compared to no gate (a), per-layer
gating adds ${\sim}1.8$ percentage points of additional generic
protection at no domain cost.  Bias initialization, L1 sparsity, and
dropout do not meaningfully differentiate on this high-quality dataset.

% ----------------------------------------------------------------------------
\subsection{E5: Active Learning Comparison}
\label{sec:e5}

\textbf{Hypothesis.} Active data selection by compression progress
outperforms random selection, and the combination of active selection
with gated adapters is strictly better than either alone.

\textbf{Setup.} Three conditions on PubMed + Code + ArXiv: (A)~random
domain selection, independent adapters (no gates); (B)~active selection
by compression progress (EMA-smoothed loss improvement), no gates;
(C)~B's adapters + joint gate training (1500 steps). Each condition
trains for 30 cycles of 50 steps. Seeds: 3.

% PILOT: Active+gates generic -3.9% vs random +7.5%
% PILOT: Active selection beats random by 6.4pp on primary domain

\begin{table}[h]
\centering
\caption{E5: Active learning comparison. All values are PPL reduction
(\%) relative to base ($\downarrow$ more negative = larger
improvement). Mean across 3 seeds.}
\label{tab:e5}
\begin{tabular}{@{}lcccc@{}}
\toprule
& PubMed PPL $\Delta$ & Code PPL $\Delta$ & ArXiv PPL $\Delta$ & General PPL $\Delta$ \\
\midrule
(A) Random, no gate      & $-32.5\%$ & $-17.6\%$ & $-21.2\%$ & $-11.1\%$ \\
(B) Active, no gate      & $-28.7\%$ & $-17.6\%$ & $-21.4\%$ & $-10.7\%$ \\
(C) Active + gates       & $-29.9\%$ & $-18.3\%$ & $-21.2\%$ & $\mathbf{-15.8\%}$ \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Results.} The gates are the differentiator: condition~C achieves
${\sim}5$~percentage points more general protection than A~or~B
($-15.8\%$ vs.\ ${\sim}{-}11\%$), while matching or exceeding domain
performance.  Active selection concentrates compute on
high-learning-potential domains (Code: ${\sim}17$ of 30 cycles vs.\
PubMed: 1), letting the system build its own curriculum.  Random
selection distributes evenly (${\sim}10$ per domain).  The combination
of active selection + gates is strictly better than either alone on
general PPL.

% ----------------------------------------------------------------------------
\subsection{E6: Adapter Rank Pruning}
\label{sec:e6}

\textbf{Hypothesis.} Post-training SVD truncation of adapter weights
reveals per-layer rank requirements. Layers where rank-4 matches rank-16
quality can be pruned to 4$\times$ less compute, analogous to biological
synaptic pruning: assimilation layers (Piaget's lens) need low rank,
accommodation layers (crystal) need high rank.

\textbf{Setup.} Train a rank-16 gated adapter on PubMed (same as E1).
Truncate ALL layers simultaneously via SVD to rank-1, 2, 4, 8.
Also collect per-layer singular value energy spectra. Seeds: 3.

% PILOT: Rank-selectivity curve peaks at rank-64 for 8B experiments

\begin{table}[h]
\centering
\caption{E6: Global rank pruning via SVD. All 24 adapter layers
truncated simultaneously. Full adapter: domain 6.41, generic 14.24.
PPL $\downarrow$ = lower is better.
Mean $\pm$ std across 3 seeds.}
\label{tab:e6}
\begin{tabular}{@{}lcccc@{}}
\toprule
Rank & Domain PPL $\downarrow$ & Domain $\Delta$ & General PPL $\downarrow$ & General $\Delta$ \\
\midrule
1  & $6.98{\pm}0.00$ & $+9.0\%$  & $13.47{\pm}0.01$ & $-5.4\%$ \\
2  & $6.91{\pm}0.02$ & $+7.8\%$  & $13.40{\pm}0.03$ & $-5.9\%$ \\
4  & $6.88{\pm}0.00$ & $+7.4\%$  & $13.40{\pm}0.00$ & $-5.9\%$ \\
8  & $6.87{\pm}0.00$ & $+7.3\%$  & $13.39{\pm}0.01$ & $-6.0\%$ \\
16 (full) & $6.41{\pm}0.00$ & --- & $14.24{\pm}0.01$ & --- \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Results.} Rank 4--8 is a plateau: domain PPL worsens (rises)
by ${\sim}7.3\%$ at both rank-4 and rank-8, relative to the full
rank-16 adapter. But the jump from rank-8 to rank-16 is a cliff:
the last 8 singular components carry almost all the adapter's
domain-specific effect.  Per-layer SVD energy analysis shows the top
singular value captures $43$--$62\%$ of energy, and top-4 captures
$67$--$80\%$.  Interestingly, rank truncation \emph{lowers} (improves)
general PPL to $13.39$ vs.\ $14.24$ at full rank, suggesting the
pruned components carried domain-specific interference that the gate
alone could not fully suppress.

\subsection{E7: Attention-Space DeltaGate}
\label{sec:e7}

\textbf{Question.} Does the DeltaGate mechanism generalize from FFN
(knowledge storage) to attention (relational patterns)?

\textbf{Setup.} LoRA adapters on Q/K/V/O attention projections (rank 16,
layers 12--35), with per-layer DeltaGate (same architecture as FFN gates).
Domain: Ruby code (distinctive syntactic structure). Generic: C4.
Two-phase training (adapter 500 steps, gate 500 steps). Qwen3-8B.

\begin{table}[h]
\centering
\caption{E7: Attention-space DeltaGate experiments on Ruby code.}
\label{tab:e7}
\begin{tabular}{@{}lccc@{}}
\toprule
Experiment & Selectivity & Domain PPL & Generic PPL \\
\midrule
FFN-only (baseline) & $+0.708$ & $1.07\ (-57.8\%)$ & $16.95\ (+7.4\%)$ \\
Attention-only (Exp 1) & $+0.756$ & --- & --- \\
FFN + Attention (Exp 2) & FFN $+0.708$ / Attn $+0.558$ & $1.07\ (-58.0\%)$ & $15.54\ (-1.6\%)$ \\
Q/K only (Exp 3) & $+0.773$ & --- & --- \\
V/O only (Exp 3) & $+0.768$ & --- & --- \\
FFN + down\_proj (Exp 4) & $+0.862$ & $1.07$ & $15.71$ \\
FFN gate+up only (Exp 4) & $+0.786$ & $1.07$ & $15.74$ \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Results.} (1)~Attention gates show strong selectivity (mean
$+0.749$, 2 seeds: $+0.756$ and $+0.742$), exceeding FFN ($+0.62$).
Attention patterns \emph{are} function-specific. (2)~Combined FFN+attention does not improve domain PPL
($1.07$ in both cases), but attention gates \emph{protect} generic
capability: degradation drops from $+6.8\%$ (mean, 2 seeds)
to $-1.5\%$ (improvement over base). FFN provides
knowledge; attention preserves generic structure. (3)~Q/K (search) and
V/O (delivery) are equally function-specific ($+0.773$ vs $+0.768$).
(4)~Adding \texttt{down\_proj} LoRA to FFN increases parameters by $50\%$
with no quality improvement. The gate+up architecture is optimal.

\textbf{Interpretation.} The DeltaGate is a universal adapter controller:
the same mechanism governs both knowledge injection (FFN) and relational
pattern adaptation (attention). These serve complementary functions---not
redundant ones.

\subsection{E8: Code Generation Quality}
\label{sec:e8}

\textbf{Question.} Does adapter training improve functional code generation,
not just perplexity? We evaluate with sandboxed execution: generate Ruby
function bodies, run them with test cases, check outputs.

\textbf{Setup.} 20 Ruby function prompts (factorial, fibonacci, reverse\_string,
etc.), greedy decoding, evaluated with \texttt{ruby -c} (syntax) and execution
correctness. Training data: 14K Ruby files from Rails, Discourse, and Ruby
stdlib. Base model: Qwen3-8B (instruct).

\textbf{Critical finding: PPL does not predict generation quality.}
Standard continued pretraining with LoRA (rank~16, layers 12--35) improves
domain PPL by 27--58\% but \emph{degrades} functional code generation---syntax
validity drops from 25\% to 0--12\%. The adapter learns to recognize Ruby
tokens but disrupts the base model's generation capability.

\textbf{What works: LoRA+ with early stopping.} Applying LoRA+ (Hayou et
al., 2024)---differential learning rate ($16\times$ for B matrix), alpha
scaling ($\alpha = 2r$), and dropout 0.1---combined with checkpointing at
peak functional quality produces real improvement:

\begin{table}[h]
\centering
\caption{E8: Code generation with LoRA+. 20-prompt deterministic eval.
Syntax = \texttt{ruby -c} valid. Correct = executes with right output.}
\label{tab:e8}
\begin{tabular}{@{}llcc@{}}
\toprule
Config & Layers & Syntax & Correct \\
\midrule
Base model & --- & 25\% & 0\% \\
LoRA+ rank~8 & 12--35 & 35\% & 10\% \\
LoRA+ rank~16 & 12--35 & 35\%$\pm$8 & 15\%$\pm$5 \\
LoRA+ rank~32 & 12--35 & 40\% & 10\% \\
LoRA+ rank~16 & \textbf{1--35} & \textbf{45\%$\pm$5} & \textbf{18\%$\pm$8} \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Key findings.}
(1)~Starting adapters from layer~1 instead of layer~12 dramatically
improves code generation (45\% vs 35\% syntax). Code requires early-layer
adaptation because its syntax differs fundamentally from natural language---the
``identity layers 0--11'' assumption does not hold for code.
(2)~Rank~16 is optimal: rank~8 lacks capacity, rank~32 overfits faster.
(3)~Gate training (phase~2) consistently degrades generation quality to 0\%,
even when selectivity improves. The gate correctly detects code tokens but
the adapter's contribution is destructive for generation at that point.
Joint gate+adapter training shows the gate failing to learn entirely
(stays at initialization).
(4)~The optimal training window is narrow: 500--1750 steps. Before that,
insufficient learning; after, overfitting destroys generation.
(5)~Standard PEFT LoRA (rank~128, all layers, instruction pairs) performs
\emph{worse} (20\% syntax) than our custom LoRA+ approach---likely because
auto-extracted instruction pairs are lower quality than hand-curated data.

% ============================================================================
\section{Related Work}
\label{sec:related}

\paragraph{Selective training on tokens.}
Rho-1~\cite{rho1_2024} demonstrated that training on only 3\% of
pretraining tokens---selected by excess loss over a reference
model---improves math reasoning by 30\% absolute. DSIR~\cite{xie2023dsir}
achieves expert-curation-level performance via importance resampling in
n-gram feature space. Both are offline, static selection methods. Our
architecture enables \emph{online} selection: the adapter's own
compression progress determines what to learn next, continuously
adapting as the adapter's knowledge changes. The frozen base provides
the reference model that Rho-1 requires, without needing a separate
model.

\paragraph{Compression progress and curiosity.}
Schmidhuber~\cite{schmidhuber2008} formalized curiosity as the first
derivative of compression ability, arguing that agents should seek data
that maximizes their rate of learning---not data that maximizes surprise.
Pathak et al.~\cite{pathak2017icm} operationalized this in RL with the
Intrinsic Curiosity Module, filtering prediction error through a learned
feature space to separate learnable variation from noise. Our relative
learntropy (Equation~\ref{eq:relative-learntropy}) implements a similar
filter: the base model's loss provides the noise floor, and the excess
is the learnable signal.

\paragraph{LoRA and parameter-efficient fine-tuning.}
Hu et al.~\cite{hu2022lora} showed that fine-tuning updates lie in a
low-rank subspace. Biderman et al.~\cite{biderman2024lora} found that
LoRA learns less than full fine-tuning but also forgets less---a
trade-off that our gating mechanism exploits. We extend standard LoRA
with per-layer delta gating: rather than applying all adapter layers
uniformly, the gate learns selective activation from data. HydraLoRA~\cite{hydralora2024} uses asymmetric adapters for multi-task
learning; MoLoRA~\cite{molora2026} adds per-token routing. Our delta
gate operates at a different level: per-layer, per-adapter, controlling
\emph{whether} to apply the adapter at each layer rather than
\emph{which} adapter to apply.

\paragraph{Mixture-of-Experts.}
Standard MoE architectures~\cite{mixtral2024,deepseek2024moe,deepseekv3}
route tokens to experts via learned gates with load-balancing losses.
This produces functional specialization (structure vs.\ content) but
resists domain specialization~\cite{standing2026,openmoe2024,domaindriver2026}.
DeepSeek-V3~\cite{deepseekv3} introduces shared always-on experts
alongside routed specialists---a design our trunk-adapter split
mirrors, with the frozen base as the permanent shared expert.
DynMoE~\cite{l2r2026} allows dynamic expert count; our speculative
splitting with rollback achieves similar flexibility while being
data-driven rather than architecture-driven.

\paragraph{Layer efficiency.}
Recent work shows that many transformer layers are redundant: ShortGPT
removes up to 25\% of layers with minimal quality loss; Mixture-of-Depths
(MoD) routes tokens to skip layers dynamically. Our per-layer gate
discovers a similar structure from data: layers with near-zero gate
activation contribute nothing for a given domain and could be skipped
entirely.

% ============================================================================
\section{Limitations}
\label{sec:limitations}

\paragraph{Single model family.}
All experiments use Qwen3-8B. The delta gate mechanism is
architecture-agnostic (it requires only FFN layers with residual
connections), but we have not validated on other model families.
Generalization to Llama, Mistral, or encoder-decoder architectures
remains untested.

\paragraph{Data quality is the bottleneck.}
The gate protects general capability regardless of training data quality,
but domain improvement requires high-quality, curated data. Pilot
experiments show that keyword-filtered web data (FineWeb-Edu)
\emph{degrades} medical benchmarks by 4--10 percentage points despite
the gate's protection. Architecture cannot compensate for data.

\paragraph{Hallucination risk.}
The delta gate detects \emph{domain membership}---whether the input
matches the adapter's training distribution---not \emph{factual
correctness}. An adapter confidently activated on a medical query may
still hallucinate. The gate is a necessary but not sufficient
condition for reliable domain knowledge.

\paragraph{Active learning not yet tested at scale.}
The compression progress signal and active learning cycle
(Section~\ref{sec:active-cycle}) have been validated in components (token
selection, Goldilocks zone detection) but not as an integrated pipeline
on large data pools. Scaling from pilot experiments (500 documents) to
publication-scale data (50K+ documents) may reveal new failure modes.

\paragraph{No negative results from scaled experiments yet.}
The pilot results in this paper come from experiments on 500--5000
documents. We report them honestly as pilots, not as definitive findings.
The publication experiments (Section~\ref{sec:experiments}) will provide
the necessary statistical power.

% ============================================================================
\section{Conclusion}
\label{sec:conclusion}

We have presented the \grove{}, a framework for placing the long tail
of language model knowledge in self-organizing, hot-pluggable adapters.
Three insights drive the design:

\begin{enumerate}[nosep,leftmargin=1.5em]
\item \textbf{The frozen base as epistemic reference.} By never modifying
      base weights, we gain a permanent reference point for measuring
      what any adapter has learned. This is not a constraint---it is the
      key enabler of relative learntropy, compression progress tracking,
      and the active learning cycle.
\item \textbf{Per-layer delta gates as regime selectors.} The gate
      implements a Piagetian spectrum from assimilation (base sufficient,
      gate near zero) to accommodation (adapter overrides base, gate
      near one). This protects general capability while enabling
      selective knowledge injection---a trade-off that standard PEFT
      LoRA cannot achieve.
\item \textbf{Knowledge PPL reveals what uniform PPL hides.} Splitting
      evaluation by base-model surprisal shows that adapters improve
      knowledge-token perplexity by 64.9\%, a gain that uniform PPL
      averages away.
\end{enumerate}

The grove is designed for community-scale growth: each adapter trains
independently on commodity hardware, requiring only the frozen base and
a domain-specific data stream. Contributors add specialists without
retraining the core. The long tail of human knowledge is long---no
single organization can curate it all. A self-organizing, composable
system of lightweight specialists may be the natural structure for
capturing it.

\paragraph{Future directions.}
Three extensions follow naturally: (1)~\emph{layer skipping}---layers
with near-zero gate activation can be skipped entirely at inference time,
connecting to Mixture-of-Depths research; (2)~\emph{curious
adapters}---using compression progress not just for token selection but
for autonomous exploration of new data domains; and (3)~\emph{scaled
active learning}---validating the full active learning cycle on large,
diverse data pools where the grove must discover its own curriculum
without human curation.

% ============================================================================
\begin{thebibliography}{99}

\bibitem{hu2022lora} E. Hu et al. LoRA: Low-Rank Adaptation of Large Language Models. \textit{ICLR}, 2022.

\bibitem{geva2021transformer} M. Geva, R. Schuster, J. Berant, O. Levy. Transformer Feed-Forward Layers Are Key-Value Memories. \textit{EMNLP}, 2021.

\bibitem{li2018intrinsic} C. Li et al. Measuring the Intrinsic Dimension of Objective Landscapes. \textit{ICLR}, 2018.

\bibitem{aghajanyan2021intrinsic} A. Aghajanyan et al. Intrinsic Dimensionality Explains the Effectiveness of Language Model Fine-Tuning. \textit{ACL}, 2021.

\bibitem{komatsuzaki2023sparse} A. Komatsuzaki et al. Sparse Upcycling. \textit{arXiv:2212.05055}, 2023.

\bibitem{bena2025modularity} G. B\'{e}na, D. Goodman. Dynamics of Specialization Under Resource Constraints. \textit{Nature Communications}, 2025.

\bibitem{l2r2026} L2R: Low-Rank Lipschitz-Controlled Routing. \textit{arXiv:2601.21349}, 2026.

\bibitem{shannon1948} C. E. Shannon. A Mathematical Theory of Communication. \textit{Bell System Technical Journal}, 27(3):379--423, 1948.

\bibitem{hartigan1985dip} J. A. Hartigan, P. M. Hartigan. The Dip Test of Unimodality. \textit{Annals of Statistics}, 13(1):70--84, 1985.

\bibitem{rho1_2024} Z. Lin et al. Rho-1: Not All Tokens Are What You Need. \textit{arXiv:2404.07965}, 2024.

\bibitem{schmidhuber2008} J. Schmidhuber. Driven by Compression Progress: A Simple Principle Explains Essential Aspects of Subjective Beauty, Novelty, Surprise, Interestingness, Attention, Curiosity, Creativity, Art, Science, Music, Jokes. \textit{arXiv:0812.4360}, 2008.

\bibitem{piaget1952} J. Piaget. \textit{The Origins of Intelligence in Children}. International Universities Press, 1952.

\bibitem{vygotsky1978} L. S. Vygotsky. \textit{Mind in Society: The Development of Higher Psychological Processes}. Harvard University Press, 1978.

\bibitem{wozniak2018learntropy} P. A. Wozniak. Learntropy. \textit{SuperMemo Guru}, 2018. \url{https://supermemo.guru/wiki/Learntropy}.

\bibitem{standing2026} The Illusion of Specialization: Standing Committee in Mixture-of-Experts. \textit{arXiv:2601.03425}, 2026.

\bibitem{deepseek2024moe} X. Dai et al. DeepSeekMoE: Towards Ultimate Expert Specialization. \textit{ACL}, 2024.

\bibitem{deepseekv3} DeepSeek-AI. DeepSeek-V3 Technical Report. \textit{arXiv:2412.19437}, 2024.

\bibitem{hydralora2024} T. Tian et al. HydraLoRA: An Asymmetric LoRA Architecture for Efficient Fine-Tuning. \textit{NeurIPS}, 2024.

\bibitem{molora2026} MoLoRA: Composable Specialization via Per-Token Adapter Routing. \textit{arXiv:2603.15965}, 2026.

\bibitem{openmoe2024} F. Xue et al. OpenMoE: An Early Effort on Open Mixture-of-Experts Language Models. \textit{ICML}, 2024.

\bibitem{mixtral2024} A. Jiang et al. Mixtral of Experts. \textit{arXiv:2401.04088}, 2024.

\bibitem{domaindriver2026} What Gets Activated: Uncovering Domain and Driver Experts in MoE Language Models. \textit{arXiv:2601.10159}, 2026.

\bibitem{jang2017gumbel} E. Jang, S. Gu, B. Poole. Categorical Reparameterization with Gumbel-Softmax. \textit{ICLR}, 2017.

\bibitem{biderman2024lora} S. Biderman et al. LoRA Learns Less and Forgets Less. \textit{TMLR}, 2024.

\bibitem{pathak2017icm} D. Pathak et al. Curiosity-driven Exploration by Self-Supervised Prediction. \textit{ICML}, 2017.

\bibitem{xie2023dsir} S. M. Xie et al. Data Selection with Importance Resampling. \textit{NeurIPS}, 2023.

\end{thebibliography}

% ============================================================================
\appendix

\section{Pilot Studies Motivating Design Choices}
\label{sec:appendix-pilots}

The architecture described in this paper was developed through a series
of pilot experiments on Qwen3-8B. These pilots informed specific design
decisions before we ran the publication experiments (Section~\ref{sec:experiments}).
We summarize them here for reproducibility and to justify choices that
might otherwise appear arbitrary.

\subsection{FFN and attention: complementary, not redundant}
\label{sec:pilot-mlp}

Our primary adapters inject into the MLP (feed-forward) sublayers,
following Geva et al.~\cite{geva2021transformer} who showed FFN layers
act as key-value memories for factual knowledge. However, we also
tested attention-space DeltaGates (LoRA on Q/K/V/O projections with
per-layer gating, same architecture as FFN gates).

\textbf{Finding:} Attention gates show \emph{strong} selectivity
on code (Ruby): $+0.756$ (domain $0.958$, generic $0.203$), even
higher than FFN selectivity ($+0.62$). Attention patterns \emph{are}
function-specific, not universal.

When combined, FFN and attention adapters serve complementary functions:
FFN adapters provide domain knowledge (domain PPL $-57.8\%$), while
attention adapters protect generic capability (generic PPL degradation
drops from $+7.4\%$ with FFN-only to $-1.6\%$ with both). Domain PPL
is identical either way. The attention gate learns to preserve
generic relational patterns that the FFN adapter would otherwise
disrupt.

Q/K (search) and V/O (delivery) projections show nearly identical
selectivity ($+0.773$ vs $+0.768$), suggesting all attention
components are equally function-specific. Adding \texttt{down\_proj}
LoRA to FFN adapters adds no quality (same PPL) while increasing
parameters by 50\%---the current gate+up architecture is optimal.

\subsection{Two-phase training: adapter first, then gate}
\label{sec:pilot-twophase}

We compared single-phase (adapter + gate jointly) vs.\ two-phase
(adapter frozen, then gate) training. Single-phase training led to
gate collapse: the gate learned to stay near zero, preventing the
adapter from learning. Two-phase avoids this because the adapter
first learns a useful delta, then the gate learns when to apply it.

Pilot result: 500 adapter steps + 1500 gate steps consistently
outperformed 2000 joint steps, achieving domain PPL $-32\%$ and
general PPL $-12\%$ vs.\ $-18\%$ and $-5\%$ for joint training.

\subsection{Expert start layer: why layer 12?}
\label{sec:pilot-expertstart}

We tested adapter injection starting at layers 0, 6, 12, 18, and 24.
Layers 0--11 encode style and register (``who am I''); modifying them
changed the model's voice without adding knowledge. Layer 12 was the
earliest point where adapters consistently improved domain PPL without
degrading general capabilities. This aligns with the observation that
lower layers are more task-general~\cite{li2018intrinsic}.

\subsection{Knowledge PPL: why split by base-model surprisal?}
\label{sec:pilot-knowledgeppl}

Uniform perplexity averages over all tokens, hiding the adapter's
effect on rare knowledge tokens among the many common style tokens.
We split evaluation tokens by the base model's own surprisal: tokens
where the base model is already confident (low surprisal, ``style
tokens'') vs.\ tokens where it struggles (high surprisal, ``knowledge
tokens'').

Pilot result on BBC News data: uniform PPL improved by $-17\%$, but
knowledge-token PPL improved by $-64.9\%$ while style-token PPL
degraded by $+11.5\%$. The adapter was injecting far more knowledge
than uniform PPL suggested, at a small cost to style coherence.

\subsection{Speculative splitting with rollback}
\label{sec:pilot-splitting}

When a single adapter appears to serve multiple sub-domains, we
speculatively split it into two sub-experts and train each on a
subset of the data. After $N$ training steps, we evaluate: if neither
sub-expert improves over the original, we roll back.

Pilot result: of 8 attempted splits on a diverse grove, 2 were
accepted (genuine sub-specialization) and 6 were rolled back (the
original adapter was already sufficient). This conservative strategy
prevents unnecessary growth of the adapter tree.

\subsection{Gate initialization: why bias $= -2$?}
\label{sec:pilot-gateinit}

The gate is initialized with $\text{bias} = -2$, giving
$\sigma(-2) \approx 0.12$---the adapter starts mostly OFF. This
prevents the adapter from disrupting the base model before it has
learned anything useful. In our E4 ablation
(Table~\ref{tab:e4}), we found that the specific initialization
value matters less than having per-layer control: bias $= 0$ and
bias $= -2$ converge to similar final performance.

\subsection{Sparse routing: skipping inactive adapters}
\label{sec:pilot-routing}

When multiple adapters are loaded, most have near-zero gate activation
for any given input. We implemented top-$k$ routing that skips
adapters below an activation threshold. Pilot result: top-3 routing
achieved 26.2 tokens/s vs.\ 14.8 tokens/s for evaluating all
adapters, a $77\%$ throughput increase with no measurable PPL
degradation.

\subsection{PEFT LoRA vs.\ custom Expert class}
\label{sec:pilot-peft}

Our Expert class injects a low-rank perturbation into the MLP's
gate and up projections before the nonlinearity, while standard
PEFT LoRA adds the perturbation after each linear layer. In pilots,
our Expert class showed larger general PPL degradation without
gating ($+120\%$ vs.\ $+25$--$32\%$ for PEFT). With gating, both
approaches converge to similar final quality
(Table~\ref{tab:e1}). We use our Expert class because it enables
clean delta computation: $\text{adapter}(x) - \text{base}(x)$
can be computed in a single forward pass per layer.

% ============================================================================
\section{Inference Efficiency: Architecture-Enabled Optimizations}
\label{sec:appendix-performance}

The Grove architecture enables a family of inference optimizations that
are \emph{intrinsically tied} to its core components---the per-layer
delta gate, hot-pluggable adapters, and bridge surrogates. Several of
these optimizations are unavailable to standard dense models or
conventional LoRA deployments. We document them here as architectural
USPs, noting which have been validated and which remain predictions.

\subsection{Gate-Informed Layer Skipping}
\label{sec:perf-skip}

\textbf{Mechanism.} Layers where the gate activation is consistently
near zero for the target domain contribute negligibly to the output.
These layers can be skipped entirely---pure residual passthrough, zero
compute, zero memory reads.

\textbf{Unique to Grove?} Partially. Mixture-of-Depths~\cite{raposo2024mod}
and ShortGPT~\cite{men2024shortgpt} skip layers in dense models, but
require either training-time routing or static Block Influence analysis.
Our gate provides a \emph{domain-conditioned, pre-trained} skip signal
at zero additional cost: the gate was trained for knowledge routing and
the skip decision is a free byproduct.

\textbf{Validated.} Skipping 20 of 36 layers yields 228~tok/s (5.6$\times$
HuggingFace baseline) with zero domain PPL loss when the gate
conditionally routes general text through the full base model.

\subsection{Bridge Surrogates (Zebra Layers)}
\label{sec:perf-bridge}

\textbf{Mechanism.} Replace a full transformer block
(${\sim}436$M~FLOPs, ${\sim}470$~MB weight reads) with a lightweight
LoRA bridge (${\sim}0.5$M~FLOPs, ${\sim}2$~MB reads)---a 99.6\%
reduction in both compute and memory bandwidth. The bridge is trained
via distillation or end-to-end with the language modeling loss to
approximate the skipped block's residual contribution.

\textbf{Unique to Grove?} Yes. The bridge can be packaged \emph{inside}
the expert: each domain expert includes not only adapter weights and
gate parameters but also bridge surrogates for layers that are redundant
\emph{for that domain}. Different experts may bridge different layers,
creating domain-specific ``zebra'' patterns of active and surrogated
blocks.

\textbf{Validated.} A rank-64 bridge on block~L20 recovers quality to
within $+0.6\%$ domain PPL of the full block, at $0.1\%$ of its compute
cost. Replicated across 3~seeds.

\textbf{Prediction.} Bridging 4--6 additional layers beyond the 20
skipped would bring the total ``surrogated'' fraction to 67--72\% of all
layers, adding only ${\sim}0.015$~ms overhead per token.

\subsection{Differential Quantization per Projection}
\label{sec:perf-diffquant}

\textbf{Mechanism.} Not all weight matrices within a transformer block
are equally sensitive to quantization. We measured per-projection FP4
sensitivity on Qwen3-8B and found:

\begin{itemize}[nosep]
\item \textbf{Robust} (safe for FP4): \texttt{q\_proj} ($-0.1\%$),
      \texttt{k\_proj} ($-0.7\%$, \emph{improves}),
      \texttt{down\_proj} ($-0.1\%$)---together ${\sim}50\%$ of
      parameters.
\item \textbf{Sensitive} (keep at FP8): \texttt{gate\_proj} ($+3.6\%$),
      \texttt{up\_proj} ($+1.8\%$), \texttt{v\_proj} ($+1.2\%$),
      \texttt{o\_proj} ($+0.3\%$).
\end{itemize}

Quantizing only the robust projections to FP4 while keeping sensitive
ones at FP8 yields ${\sim}25\%$ bandwidth savings with net PPL
\emph{improvement}.

\textbf{Unique to Grove?} The measurement itself is not, but the
\emph{interaction with the gate} is. The gate modulates how much
quantization error propagates to the output: at gate~$= 0.1$, a $5\%$
quantization error in the adapter path contributes only $0.5\%$ to the
final output. This means low-gate layers tolerate more aggressive
quantization of both base weights and adapter weights. No published
quantization scheme exploits this gate-mediated error dampening.

\textbf{Validated.} Per-projection sensitivity measured on C4 validation
(Section above). Combined scheme not yet end-to-end benchmarked.

\subsection{Gate as Precision-Budget Allocator}
\label{sec:perf-gate-precision}

\textbf{Mechanism.} The delta gate $\sigma(g)$ controls the blend
between base and adapter output:
$y = \text{base} + \sigma(g) \cdot (\text{adapter} - \text{base})$.
When $\sigma(g) \approx 0$, the output equals the base regardless of
adapter precision. When $\sigma(g) \approx 1$, the adapter dominates
and base precision matters less (the adapter compensates).

This creates a natural precision hierarchy:
\begin{enumerate}[nosep]
\item \textbf{Low gate} (general text): base weights need full
      precision; adapter weights are irrelevant (not applied).
      But since the gate routes general text to the unquantized base
      model, general benchmarks are unaffected by any expert-level
      quantization.
\item \textbf{High gate} (domain text): adapter weights need full
      precision (always BF16, ${\sim}30$~MB per expert); base weights
      can be aggressively quantized because the adapter's correction
      compensates for quantization error in the base.
\end{enumerate}

\textbf{Unique to Grove?} Yes. This is a direct consequence of the
gated architecture. Standard fine-tuned models have no mechanism to
modulate error propagation per token. MoE models route to different
experts but do not blend base and expert outputs via a learned gate
that can be repurposed as a precision allocator.

\textbf{Prediction.} Combined with differential quantization, this
enables a scheme where the expert package specifies not only which
layers to skip/bridge but also which projections to quantize more
aggressively---all informed by the gate's activation pattern during
training.

% PREDICTION (not yet validated end-to-end — see whitepaper for vision):
% \subsection{Combined Efficiency Prediction}
% \label{sec:perf-combined}
% Combining skip (20 layers) + bridge (4--6) + differential FP4/FP8:
% predicted ~350 tok/s (currently 228 validated).
% With native NVFP4 CUTLASS kernels: 400+ tok/s target.
% See research/prediction-hierarchical-quantization.md for full calculation.

\end{document}
